Pre-Cambrian origin of envelope-carrying retrotransposons in metazoans

Version 1

Abstract

Retrotransposons or endogenous retroviruses (ERVs) essentially carry open reading frames of gag and pol, which are utilized to selfishly replicate themselves in the host germline genome. One rare example of ERVs that additionally carry envelope genes, are Ty3/gypsy errantiviruses in Drosophila. Though they are structurally analogous to retroviruses, their evolutionary relationship as well as the origin of envelope genes remain elusive. We systematically searched for intact envelope-containing ERVs that are homologous to Ty3/gypsy in invertebrate metazoan genomes and found that they are widespread across taxa including ancient animals such as cnidarians, ctenophores and tunicates. Most elements are multiplied in their respective genomes, implying recent germline expansion. envelope genes are classified into those that resemble glycoprotein F from paramyxoviruses and glycoprotein B from herpesviruses, and both types are equally abundant and widespread. Phylogenetic and structural analyses revealed that envelope genes have largely diverged with pol genes as well as with the host organisms throughout their evolutionary history and recombined infrequently, suggesting that the envelope acquisition to ERVs is ancient and likely dates to before the split of bilaterian and non-bilaterian animals in Pre-Cambrian era.

Introduction

Long terminal repeat (LTR) retrotransposons are one of the most abundant classes of transposons that are ubiquitously present in eukaryotic genomes. They replicate by reverse-transcribing the messenger RNA and inserting the DNA copy into the genome. They essentially carry two open reading frames (ORFs), namely, gag and pol. gag encodes for a structural protein that packages RNA into capsids while pol encodes a multi-functional protein that contains the reserve transcriptase, RNase H and integrase domains. Gag and pol ORFs are flanked by 5’ and 3’ LTRs that act as a promoter for transcription as well as a primer for reverse transcription. Although intact copies can continue to replicate, individual insertions acquire mutations and degrade overtime once they become inactive. Some retrotransposon remnants remain in the genome to provide functional ORFs and regulatory sequences of host gene expression.

It is widely accepted that retroviruses are evolutionarily sister to LTR retrotransposons among other retroelements (Xiong and Eickbush 1988; Xiong and Eickbush 1990). In addition to pol and gag, retroviruses additionally carry the envelope gene, which encodes a membrane fusion protein essential for infectivity. Although they typically horizontally transmit between individual animals, they can occasionally infect the germline cells of the host, vertically transmit thereafter, and become known as endogenous retroviruses (ERVs). A well-studied example of such endogenization events is found in Koala, in which nearly identical copies of an exogenous lymphoma-causing retrovirus are found in the genomes of geographically distinct Koala populations (Tarlinton et al. 2006; Zheng et al. 2020). Although individual ERV insertions are subject to genetic drift and degrade overtime after they acquire mutations, some insertions remain infectious and continue to make new insertions in the germline genome (Ribet et al. 2008; Miyazawa et al. 2010; Anai et al. 2012; Verneret et al. 2025). However, it is unclear how long such infectious ERVs vertically transmit in a long evolutionary time. When homologous ERVs are found in distantly related species, it can be either seen that they have been inherited from their common ancestor or recently transmitted from one species to another through horizontal transfer. The mechanism of such horizontal transfer events is even less clear although it is often hypothesised that exogenous retroviruses that are borne by one species infect the other species that shares the habitat (Hayward et al. 2020; Simpson et al. 2022; Mottaghinia et al. 2024).

envelope-carrying ERVs are also found in invertebrate genomes, to which no exogenous retroviruses are currently known. Some gypsy ERVs or better known as gypsy LTR retrotransposons in Drosophila carry an envelope gene (Song et al. 1994; Pelisson et al. 2002). These so-called errantiviruses behave like bona fide retroviruses, making virus-like particles in the ovarian somatic cells and infecting the germline cells (Song et al. 1997; Leblanc et al. 2000). Therefore, errantiviruses replicate within the genome of the same animal as other retrotransposons do but through an intra-gonadal infection mechanism. Errantiviruses are widespread in Drosophila genomes and form a monophyletic group among other Drosophila gypsy elements based on the phylogeny of the pol ORF, suggesting that a subtype of gypsy acquired the envelope gene in the common ancestor of Drosophila and has been vertically transmitting thereafter (Bargues and Lerat 2017). It is postulated that the ancient gypsy retrotransposon in Drosophila acquired the envelope gene from an insect-specific DNA virus, baculovirus, through recombination (Malik et al. 2000; Rohrmann and Karplus 2001). However, Ty3/gypsy LTR retrotransposons with envelope genes have been also sporadically documented in non-insect animals, such as in nematodes, rotifers, and tunicates, obscuring their true origin (Volff et al. 2004; Rodriguez et al. 2017; Sacco et al. 2022).

Here we comprehensively searched for envelope-carrying Ty3/gypsy retrotransposons in invertebrate metazoan genomes. We found elements with uninterrupted ORFs flanked by putative LTR sequences in nearly every major phylum, including ancient animals such as cnidarians, ctenophores, and tunicates. Furthermore, most of these elements are present in multiple copies in their respective genomes, indicating recent transposition. Phylogenetic and structural analyses of the pol and envelope genes show that they have co-evolved with their host species for a long evolutionary time. Therefore, we propose that the acquisition of the envelope gene to retrotransposons occurred anciently in metazoans, and they have been vertically transmitting since pre-Cambrian era.

Results

The approach to study the phylogeny and evolution of LTR retrotransposons, is conventionally to mine the genome for sequences which share homology to retroviral gag, pol, or envelope (env) genes. Prior studies on ERVs and retrotransposons have assembled trees based on reverse transcriptase (RT) or Integrase (IN) from pol to obtain insights (Xiong and Eickbush 1990; Wang and Han 2021). One can also investigate synteny of loci between different hosts, estimating an age of origin for those elements (Bonnaud et al. 2005; Ueda et al. 2020).

This study aims to investigate the evolution of env genes acquired by active Ty3/gypsy retrotransposons, and to explore structural variations within those elements. The conventional approach, therefore, needs to be altered to approach this. We only consider intact copies of env-containing retrotransposons; those that carry uninterrupted open reading frames (ORFs) for all three retroviral proteins (Fig 1A). We directed our search within a diverse array of hosts, and where possible, focused on such elements that are multiplied in the genome to understand relationships of currently active elements. This approach enables us to study structure and phylogeny more accurately, as structural predictions will be more confident on non-degraded elements. Additionally, the study enables the discovery of new elements that are active today, potentially providing insights on more recent acquisition events. The study of syntenic loci here is inappropriate, as the genomic location of active transposons will likely never be the same in evolutionarily distant hosts. Like most studies of repeat elements, the sequences are highly diverse, and we therefore supplement phylogenetic evidence with structural predictions.

The widespread nature of intact GAG-POL-ENV elements across metazoan genomes.

(A) Open reading frame arrangement of intact errantivirus copies. (B) Summary of Errantiviruses found in Metazoan phyla with type of ENV protein indicated. (C) Copy number of intact errantivirus copies in metazoans. Column represents a genome assembly, each tile representative of annotated errantivirus, colour of tile represents copy number.

We systematically searched for homologous elements of Dipteran Ty3/gypsy errantiviruses in the referenced genomes across metazoan phyla using the annotated protein coding sequences of gag, pol and env for tBLASTn searches (see methods and Fig S1). The first tBLASTn search identified elements outside Diptera, therefore, we extended the search using the representative of newly identified elements as baits. We repeated this process and examined whether the additional baits uncovered elements that were not represented in the previous tBLASTn searches by identifying new clades in the phylogenetic tree of POL (see methods and Fig S1). After a five iteration of tBLASTn searches, we no longer discovered elements that formed a new POL phylogenetic clade (Fig S1B). These steps allowed us to obtain a comprehensive set of elements in metazoan genomes that are homologous to Dipteran Ty3/gypsy errantiviruses (Table S1, summarised in Fig 1). The majority of the newly identified genomic copies carry putative LTRs and tRNA primer binding sequences (1151 out of 1512 elements, Table S1), suggesting their recent integrations. Additionally, most of the annotated elements sit within contigs larger than 0.1 mega bases (1434 elements, Table S1), indicating their true origins in their respective genomes. We ran multiple sequence alignments using the conserved Integrase (IN) and Reverse Transcriptase (RT) domains of Pol and assembled maximum probability trees to infer phylogenetic relationships of the identified elements. ORFs of annotated Drosophila gypsy errantiviruses were also included in the trees to aid classifications. We performed a similar analysis on vertebrate genomes using known vertebrate retroviral gag, pol, and env sequences as bait in addition to Drosophila errantiviruses. We did not identify any intact ERVs in vertebrate genomes that were homologous to invertebrate elements, nor did we find elements in invertebrate genomes that predicted homology to vertebrate elements. There is therefore a clear phylogenetic separation between invertebrate elements and vertebrate ERVs, despite structural similarities. We refer to the genomic copies that are homologous to Drosophila gypsy as errantiviruses in this study. The investigation on the vertebrate ERVs will be reported elsewhere, as it within a different scope to errantiviruses.

Multiple intact elements of env-carrying Ty3/gypsy retrotransposons are found widespread across metazoan species

We found that intact copies of errantiviruses carrying all three retroviral ORFs are found not only outside of insect orders but are ubiquitously present across most major metazoan phyla. This includes species from ancient phyla such as Cnidaria, Tunicates and Ctenophora, but excludes Echinodermata and Porifera (Fig 1B and Table S1). Many carry ENV proteins that are homologous to glycoprotein F1, commonly found in paramyxoviruses, such as Nipah, Hendra and Respiratory Syncytial Viruses (McLellan et al. 2013; Dang et al. 2019), but is also carried in insect baculoviruses (and Drosophila errantiviruses) (Malik et al. 2000; Rohrmann and Karplus 2001). Other elements we identified instead carry ENV that resembles glycoprotein B (gB), famously found as part of the glycoprotein complex in Herpes Simplex Viruses (HSV). Both types of ENV glycoproteins are carried by errantiviruses found in ancient hosts such as rotifers, tunicates, cnidarians and ctenophores, as well as in hosts from insect and arthropod orders (Fig 1B). Though, errantiviruses found in Nematoda, Bryozoa and Platyhelminthes only carry HSV/gB-type ENV, and the ORFs are arranged as GAG-ENV-POL rather than GAG-POL-ENV. Most elements that were identified have multiple intact copies in their respective genomes (as defined as copies with >98% nucleotide identity across >98% of the three ORFs, see methods, Table S1), a hallmark feature of recent transposition (Cordaux and Batzer 2009; Gagnier et al. 2019), indicating that most elements that we have annotated likely behave as bona-fide retrotransposons (Fig 1C). Activity of these elements are highly varied, as the copy number can be as little as one intact copy, to hundreds of intact copies in the genome (Fig 1C, Table S1). Importantly, multi-copy ERVs are present in nearly all animal phyla where we see intact elements, suggesting that these elements are actively transposing in a wide range of metazoan species today.

The phylogenetic tree assembled on the intact errantivirus pol genes shows congruence, meaning that the tree assembled on pol RT closely reflects that of pol IN. Errantiviruses within most clades in one tree are represented in the same or closely related clades in the other (Fig S2B). Importantly, the errantiviruses found in phyla outside of insects, are in a different clade to those found within, which we call “ancient errantiviruses” and “insect errantiviruses”, respectively (Fig 2 and S2A). Notably however, few elements found in insects sit within clades of ancient errantiviruses, rather than other insect ones (Fig 2). Annotated Drosophila gypsy elements included in the tree, namely, Gypsy-ZAM, Gypsy-Gypsy/Gypsy-5_DGri and Gypsy-DM176/Idefix present themselves in different clades while they cluster together with other insect errantiviruses, in an arrangement with previous studies that have classified Drosophila gypsy elements (Bargues and Lerat 2017) (Fig 2). Though, one element from a cnidarian species Paleozoanthus reticulatus clusters together with insect elements, in a clade adjacent to DM176 and Idefix (Fig 2), its origin is unclear as the element is found on a short genomic contig (Table S1).

Phylogeny of Reverse Transcriptase (RT) from all intact metazoan errantiviruses.

Maximum likelihood tree constructed on representative pol RT proteins of intact errantiviruses (see methods) and rooted to the midpoint. The “ancient errantiviruses” and “insect errantiviruses” phylogenetic groups are highlighted in green and white respectively. Positions of known dipteran errantiviruses, Gypsy_ZAM (1), Gypsy_Gypsy5_Dgri (2), Gypsy_Gypsy (3), Gypsy_Idefix (4), and Gypsy_DM176 (5) are indicated. Host classifications of errantiviruses are annotated in colour, non-arthropod Protostomia (yellow), Deuterostomia (red), Diploblasta (orange), non-insect Arthropoda (green), Insecta (blue). One cnidarian element within the group of insect errantiviruses is marked by an asterisk.

Many of the newly identified “insect errantiviruses” sit in separate clades to the Annotated Drosophila gypsy elements (Fig 2, tree positions I3 - I6 in Fig S2A). These “insect errantivirus” clades are strongly supported by bootstrap values greater than 95 and are additionally supported by the annotations of the tRNA primer binding site (PBS) found in each element. For example, Tree position I1 within “insect errantiviruses” mainly carry tRNASer, whereas Tree position I2, consisting of only dipteran elements, predominantly use tRNALys, though few use tRNASer or tRNAPro (Fig S2A and S2C). Elements in insects within positions I4, I5 and I6 predominantly carry tRNALeu primer binding sites, in addition to Serine and other uncommon tRNAs (Fig S2C). This is mostly consistent with past works, that have identified Drosophila gypsy elements to carry either Serine or Lysine tRNA PBS (Terzian et al. 2001; Stefanov et al. 2012). Notably, most discrete clades comprised elements from the same insect orders, suggesting long term vertical transmission and divergence within related species (Fig S3).

The “ancient errantivirus” group also consists of clades strongly supported by bootstrap values greater than 95, though the tRNA PBS content within the clades is much more divergent (Fig S4C). Notably however, errantiviruses found in insects that sit within this group, in tree positions A1, A2 and A7 continue to carry predominantly tRNASer, tRNALys and tRNALeu (Fig S4B). Discrete clades are consistently made up of elements in hosts from the same, or closely related phyla (Fig S4B), for example, tree position A5 consists of elements found in Thecostraca, Annelida and Mollusca, each of which form their own discrete clade (Fig S4B). This indicates that they were either present in their common ancestor or inherited more recently within the same animal group. Given that neighbouring clades consist of elements found in related animal phyla, it is more likely that they are ancient in nature. For example, clades that consist of insect elements are flanked by clades of other arthropod elements (Fig S4B, tree position A2). Clades containing Annelida elements often cluster with or next to mollusc elements (Fig S4B, tree positions A5-A7). This congruence between clades and species is highly indicative of long-term vertical transfer, as known in non env-containing retrotransposons (Sormacheva et al. 2012; Wanner and Faulk 2021). Within larger clades containing a wider variety of host orders such as tree position A9, the separation is still evident and strongly supported clades only consist of elements from the same order of host (Fig S4B). Furthermore, clades formed on the trees using pol proteins tend to carry the same type of env (Fig 3A and S4A). For example, within A6 in the “ancient errantivirus” group, elements that carry HSV/gB-type env are in a separate clade to those carrying F-type env (Fig S4B). Though some clades appear to be a mix of both, elements with HSV/gB-type env still tend to cluster together, indicating that they have stayed with pol for long evolutionary periods of time.

env classifications of intact metazoan errantiviruses.

(A) Shown is a pol RT maximum likelihood tree of the representative errantiviruses (see methods) with host (*) and type of ENV protein (**) indicated. Colours represent non-insect animals (*) (yellow), insects (*) (blue), glycoprotein F (F) (**) (orange), and glycoprotein B (HSV/gB) (**) (purple). Clades containing hymenopteran elements in “Insect errantiviruses” that carry either F or HSV/gB env are annotated as 1 and 2. (B) Close up of clades 1 and 2 in “Insect errantiviruses”. F-type and HSV/gB-type errantiviruses that are both found in Lasioglossum lativentre are marked by asterisks.

ENV glycoproteins carried by errantiviruses are highly structured, and their divergence follows that of pol

As previously mentioned, we found that errantiviruses can carry HSV/gB-type or glycoprotein F-type env. While most HSV/gB-type env is found in the “ancient errantivirus” clade, one clade in “insect errantiviruses” exclusively consists of hymenopteran errantiviruses which carry both HSV/gB-type and F-type env genes (Fig 3A-B). Interestingly, elements with HSV/gB-type env and F-type env separate at the base of this clade while they both are found in a wide range of species superfamily within Hymenoptera, suggesting that both HSV/gB- type and F-type env have been together with their pol genes at least throughout the evolution of hymenopteran species (Fig 3B).

To further characterise the ENV proteins carried by errantiviruses, we identified the ectodomains within the env ORF, which is flanked by signal peptides and a transmembrane domain, using DeepTMHMM (https://doi.org/10.1101/2022.04.08.487609). We then analysed trimeric configurations of these ectodomains from representative errantiviruses from the phylogenetic tree using Alphafold 2 (Jumper et al. 2021).

F glycoproteins analysed are highly structured and consistently form four domains that are highly conserved across all elements (Fig 4A). They also predicted a highly conserved Furin cleavage site and hydrophobic fusion peptides in the ectodomain (Fig 4A and S5). The 3D fold of the ectodomain structures forms a “head” (domains 1-3) and a “stalk” (domain 4), strongly resembling 3D structures of F ectodomains characterised in previous studies, such as those from RSVs (McLellan et al. 2013) (Fig 4C). Most errantiviruses also have an α helix at the C terminus of the ectodomain, referred to as “Domain A” (Fig 4A and 4C). The predicted 3D trimeric structure also reveals two potential configurations of the glycoprotein, based on domain 4: either a long α helix, or two shorter helices. These long and short barrel structures resemble pre- and post-fusion structures of glycoprotein F described in prior studies (McLellan et al. 2013). These highly structured domains in the intact env ORFs in these errantiviruses demonstrates that env has remained intact and could suggest that their function also remains intact as well.

Structure and phylogeny of glycoprotein F-type ENV found in intact errantiviruses.

(A) Domain arrangement of glycoprotein F ectodomain in Errantiviruses. (B) A Sankey plot of pol RT and glycoprotein F- type ENV ectodomain maximum likelihood trees for representative errantiviruses, showing a phylogenetic congruence between them. Bootstrap values for nodes are given. The “ancient errantivirus” group is highlighted in green. (C) Monomeric and trimeric structures of the “pre-fusion configuration” and the “post fusion configuration”. Schematic cartoons of F-type ENV ectodomain domain 4 are shown in (C’) and (C’’) Respectively. Structures are taken from Diptera_Sciomyzidae_Cmar_errantivirus_19 and Branchipoda_Diplostraca_Dcar_errantivirus_3 for “pre fusion” and “post fusion” structures in (C).

To explore the phylogenetic relationships between pol and env-F, we generated multiple sequence alignments of the ectodomains and RT domains from representative elements and assembled phylogenetic trees. We found that the ENV ectodomains cluster similarly to RT in pol, and ENV proteins tend to form clades within the same or closely related species as they do for RT (Fig 4B). Importantly, the ENV ectodomains found in “ancient errantiviruses” cluster separately to those found in “insect errantiviruses”, including those insect elements that present themselves with ancient errantiviruses in the RT tree (Fig 4B). This suggests that env-F genes have stayed with pol, and the acquisition of env likely occurred anciently.

The ectodomain of errantiviruses that carry HSV/gB-type env are either flanked by signal peptides and a transmembrane domain, or by two transmembrane domains and the Furin cleavage site is not always conserved (Fig 5A). Like F, the trimeric structures of HSV/gB-type ENV in errantiviruses are also highly structured. They carry five highly conserved and distinct domains (designated as domain I – V) arranged in a hairpin configuration, that fold into a trimeric structure with a base, stalk and crown (Fig 5B). Additionally, domain I carries highly conserved “fusion loops” consisting of hydrophobic amino acid residues. These domains have also been characterized in Herpes Simplex Viruses in prior studies (Heldwein et al. 2006; Connolly et al. 2011). Notably, errantiviruses found in insect orders that carry HSV/gB-type env, only carry one of these fusion loops (Fig S6). The structure of the HSV/gB-type ENV carried by these errantiviruses are stabilized by disulphide bonds between cysteine residues, which we refer to as cysteine bridges (Lopper and Compton 2002; Vollmer et al. 2020). These bridges are highly conserved but appear to vary the most within domains I and II of the HSV/gB ectodomain (Fig 5C and S6). Additionally, in some cases, a cysteine bridge within domain I is located near the fusion loops (Fig 5C and S6). The configuration of cysteine bridges appears to be specific to phyla, though this does not seem to be reflected by their positions in the tree based on pol RT. For example, Bryozoa elements (Tree positions A3 and A9 in fig S4B) all carry configuration ‘C1c’, despite being found in different clades on the tree, even though they each cluster themselves separately from other elements (Fig S4B, 5D and S6). Similarly, errantiviruses carrying HSV/gB-type env found in insect phyla mostly have the same configuration regardless of their position in the pol RT tree. For example, insect elements within the “ancient errantivirus” group, and insect elements within the “insect errantivirus” group, all have the same cysteine bridge configuration in HSV/gB (Fig 5D and S6). There are, however, cases where more than one type of configuration can be found within the same phyla, namely Nematoda and Platyhelminthes (Fig 5D). Though, in those cases, they appear to continue to separate based on pol RT (Fig S7). For example, Nematoda elements with configuration ‘C2a’ and ‘C2c’ are found in a different clade to those with ‘C1a’ and ‘C1d’ (Fig 5D). Although, it should be noted that elements with ‘C1a’ in Nematoda are found in two different clades (Fig 5D).

Structure of glycoprotein HSV/gB-type ENV found in intact errantiviruses.

(A) Structural arrangements of Ectodomains, Transmembrane domains and signal peptides found in HSV/gB-type errantiviruses. (B) Shown are the arrangement of domains of the HSV/gB-type ENV ectodomain, and Alphafold 2-predicted 3D monomer and trimer structures of the ectodomain. The structure is taken from Hemiptera_Pemphigidae_Elan_errantivirus_1. (C) Fusion loops and conserved cysteine bridges on structure of the errantivirus HSV/gB-type ENV ectodomain. Shown are variable cysteine bridges in the errantivirus HSV/gB-type ENV ectodomains, in domain I and II (Indicated with orange bar. Proximity of cysteine bridge to fusion loops of domain I are indicated for structures within C1, C2 and C3. Structures taken from Cnidaria_Hydrozoa_Chem_errantivirus_5, Bryozoa_Gymnolaemata_Cpal_errantivirus_2, and Tunicata_Ascidiacea_Clep_errantivirus_1, for C1, C2 and C3 configurations, respectively. (D) Shown is the summary of the cysteine bridge configurations found in HSV/gB-type ENV of different metazoan orders. Positions for cysteine bridge configurations within Nematoda and Platyhelminthes shown on Simple schematic. Exact positions of all cysteine bridges can be found in Supplementary Figure S6.

As mentioned previously, the clades that form on RT mostly separate elements that carry HSV/gB-type ENV from F-type ENV within discrete clades (Fig S4B). This, with the fact that the same substructures can be found in different orders, for example in Nematoda, Bryozoa and Platyhelminthes, suggests that their origin is likely ancient. The overall specificity of certain residues to host orders further suggests that recombination events would have only occurred between elements within the same animal order. Furthermore, the overall separation of distinct cysteine bridge configurations of HSV/gB-type ENV, that are represented in the pol RT tree within ancient phyla, suggests that those recombination events occurred anciently. While insect elements are separated on the tree, ERVs with HSV/gB-type ENV tend to cluster separately from those with F-type ENV. For example, in the “insect errantivirus” group, the clade that almost exclusively carries HSV/gB-type ENV sits sister to another clade that only carries F-type ENV (Fig 3B). Not only do they separate from each other, but HSV/gB- type ENV can be found in a variety of superfamilies within Hymenoptera, which suggests their ancient origin (Fig 3B). Some species, such as a hymenopteran Lasioglossum lativentre, carry elements with HSV/gB-type, as well as those with F-type ENV (Fig 3B and Table S1). Despite being found in the genome of the same species, the pol of errantiviruses with F- and HSV/gB-type ENV do not stay together, further supporting their more ancient origin (Fig 3B). It should be noted that some clades, such as ones in tree position A2 within “ancient errantiviruses”, have elements that carry F- and HSV/gB-type ENV sitting together within the same discrete clade, suggestive of more recent recombination (Fig S4B).

The RNase H Domains within the bridge region of pol separates “insect errantiviruses” from “ancient errantiviruses”

To further characterize the errantiviruses identified in this study, we analysed the region between the IN and RT domains of Pol, referred to as the “bridge region”, for representative elements. This region is known to carry a ribonuclease H (RNase H) domain, which in vertebrate retroviruses has duplicated, and the N-terminal copy, while partially degraded, maintains the same structure and 3D fold arrangement, referred to as “tether” (Malik and Eickbush 2001). We therefore characterised structures within the bridge region using the 3D fold arrangement of the core β-sheet structure using Alphafold 2.

We found various structures within the bridge region, mostly consisting of core β-sheets resembling the 3D fold arrangement of RNase H domains but we also found other uncommon substructures (Fig 6A). Almost all errantiviruses that sit within the “insect errantivirus” group have duplicated the RNase H domain similarly to vertebrate retroviruses, with the C terminal copy presenting itself as a “pseudo” RNase H domain (Fig 6A-B). This “pseudo” RNase H domain retains the barrel-like structure made of five core β-sheets, but the order of the second and the third β-sheets is flipped (Fig 6A).

RNase H domains found in the bridge region of intact errantiviruses.

(A) Alphafold 2-predicted structures of domains found in the bridge regions of errantiviruses, with topology maps of α helices (grey) and β sheets (red). Structures are taken from Annelida_Polychaeta_Apac_errantivirus_7, Bryozoa_Gymnolaemata_Cpal_errantivirus_3, Coleoptera_Coccinellidae_Npum_errantivirus_1, Annelida_Polychaeta_Slim_errantivirus_9, Diptera_Cecidomyiidae_Smos_errantivirus_1 and Hymenoptera_Apoidea_Obic_errantivirus_2 for RNase H/Tether, ‘mini’ RNase H, ‘pseudo’ RNase H domains, “Mini” domain 2, “Pseudo” domain 2 and “Pseudo” domain 3, respectively. (B) Shown is the pol RT maximum likelihood tree of representative errantiviruses, with RNase H domain architecture in bridge region of pol highlighted. “Insect errantiviruses” with RNase H + pseudo–RNase H (blue), “insect errantiviruses” with bridge region outliers (dark blue), “ancient errantiviruses” with RNase H + mini domain (green), “ancient errantiviruses” with bridge region outliers (yellow). (C) The structure of the bridge region of yeast Schizosaccharomyces pombe Ty3 element tf1 predicted by Alphafold 2.

On the other hand, “ancient errantiviruses” carry an RNase H domain, and a C-terminal “mini” domain. This consists of three small core β-sheets and α-helix residues at its N terminus. This “mini” domain is also carried by insect elements in this group, further suggesting that those insect elements are equally ancient as the other errantiviruses in this group. Furthermore, this “mini” domain is also found in the bridge region of yeast S. pombe tf1 pol (Fig 6B-C), indicating it as a feature of highly ancient transposons.

The other structures (“Mini” domains 2, 3 and “Pseudo” domain 2) present themselves less frequently as outliers, and are restricted to specific clades, or even solitary elements (highlighted in yellow and dark blue in Fig 6B). One notable outlier is an insect element within the “ancient errantivirus” group that carries a variation of a “pseudo” RNase H domain (position 7, noted as “pseudo” domain 2 in Fig 6). This same domain can be found in a phylogenetic group in the “insect errantivirus” group (position 3 in Fig 6B). Furthermore, some elements in this “insect errantivirus” group carry a “mini” domain as well (position 3 in Fig 6B). Another is that some elements within the “insect errantivirus” group only carry a solitary RNase H domain (positions 4 and 5 in Fig 6B). It should be noted that these outliers are mostly specific to small phylogenetic groups/clades, and likely also arose anciently.

Active errantiviruses are sourced from various lineages of Ty3/gypsy retrotransposons

To further investigate the origins of env-containing errantiviruses, we assembled a phylogenetic tree from the RT of representative errantiviruses and included non-env-containing retrotransposons that fall under the broader umbrella of “Ty3/gypsy” elements. This includes transposons such as MAG and Tor, and Chromoviruses such as Maggy, which are annotated in the Gyspy DB (https://gydb.org). We expected that env-containing errantiviruses would form a monophyletic clade in the tree if they arose from a specific non-env-containing Ty3/gypsy element.

We firstly find that most non-env-containing Ty3/gypsy elements sit in the ‘ancient errantivirus’ clade, with the exception of the 412/MDG1 group - specifically the Drosophila gypsy elements BLOOD and Gypsy-28_Dan - which cluster with Drosophila and other insect errantiviruses, consistent with a previous study (Bargues and Lerat 2017) (Fig 7). Additionally, we find that the non-env-containing elements rarely regularly form discrete clades without env-containing errantiviruses. For example, Tor4b is found sister to Tunicata elements carrying F-type env, and Cer1 was found sister to a Platyhelminthes errantivirus carrying HSV/gB-type env. Interestingly, Chromoviridae, which are found in Fungi and Plants and are representative of highly ancient, widespread Ty3/gypsy elements (Gorinsek et al. 2004; Kordis 2005), are found to cluster within the tree, nested with “ancient” errantivirus clades (Fig 7). They sit within a clade that includes Tunicata errantivirus carrying HSV/gB-type env.

Trees of known Ty3/gypsy elements and newly identified Errantiviruses.

Midpoint rooted pol RT maximum likelihood tree of representative errantiviruses inclusive of non-env Ty3/gypsy pol RT. Host Order and env type given for each clade, as well as solitary elements on the tree are shown. Errantiviruses containing chromodomains are marked in yellow. The “ancient errantivirus” group is coloured in green and “insect errantiviruses” in black. Known Ty3/gypsy elements includes Gypsy Group 1 OSIRIS (Gypsy-21_DAn), Gypsy Group 1 OSVALDO (Dypsy-3_DEl), Gypsy Group 2 BICA (Gypsy-6_DFi), Gypsy Group 2 BLASTOPIA (Gypsy-32_DWil), Gypsy Group 2 MDG3 (INVADER3, MDG3_DM), Gypsy Group 2 MICROPIA (Gypsy- 7_DSe), Gypsy Group 2 SACCO (Gypsy-10_DAn), Gypsy Group 3 412/MDG1 (BLOOD, Gypsy-28_Dan), Gypsy Group 3 CHIMPO (Gypsy-7_DAn), Gypsy Group 3 17.6 (DM176, IDEFIX, ZAM), Gypsy Group 3 GYPSY (Gypsy-5_Dgri, Gypsy-Gypsy), Mag (Mag_As, MAG_Bmor), Cer (Cer1, Cer4), Tor (Tor1, Tor2, Tor4b), Athila/Tat (ATHILA0p1), and Chromoviridae (ATHILA0p1, CRM_AAM94350.1, REINA, Saccharomyces_Ty3, SCL213_Galadriel, Skipper*, Spombe_Tf1_AAA35339.1, SUSHI).

The annotated Ty3/gypsy elements present themselves in the tree, spread between and within different “ancient” errantiviruses from different hosts, with little pattern to their spread regarding host or env type. Notably, a clade of errantiviruses from Cnidaria, Mollusca and Annelida hosts all carry the chromodomain, and the clade sits basally to known chromoviruses, but sister to non-chromodomain-containing errantiviruses (Fig 7, Fig S7). We also observe outliers, where only one or two errantiviruses carry chromodomains within the whole clade. For example, Annelida_Polychaeta_Dgyr_Errantivirus_1 contains a chromodomain, but is located in a clade with non-chromodomain-containing errantiviruses (Fig S7). These observations rule out a single origin of env- containing errantiviruses and instead suggest that multiple lineages of ancient Ty3/gypsy elements contributed to the current distribution and diversity of errantiviruses.

Discussion

We found copies of env-containing Ty3/gypsy retrotransposons (errantiviruses) in the genomes of nearly every major metazoan order. They are intact, carrying uninterrupted ORFs of all three retroviral proteins and are flanked by LTRs. Additionally, most errantiviruses identified in this study have multiple copies in their respective genomes, likely indicating recent transposition. When we further characterized these elements, we found that they carry different ENV proteins; ENV resembling glycoprotein F from paramyxoviruses, and ENV resembling glycoprotein B from herpesviruses. Despite great divergences in the primary sequence, the structural domains of these proteins are highly conserved. We therefore propose that the term ‘errantivirus’ should be used to refer to such env-containing Ty3/gypsy elements found outside Drosophila as well.

We found that these errantiviruses are highly ancient, being found in animals such as cnidarians, ctenophores and tunicates – lineages that diverged early in metazoan evolution. We identified an “ancient errantivirus” group, which sits separately to elements found in insects (Fig 2). These ancient elements are also structurally distinct, with the bridge region between RT and IN of pol resembling that of yeast Ty3 retrotransposons, rather than that of insects. Additionally, pol genes of non-env-containing Ty3/gypsy elements from diverse eukaryotic taxa form clades with the “ancient errantivirus” group (Fig 7). Although these patterns are consistent with their ancient nature, the specific origins of env-containing errantiviruses in metazoans remain elusive. Errantiviruses from the same or closely related phyla tend to cluster together in the pol RT phylogenetic tree (Fig S4). However, the pattern becomes convoluted when highly ancient errantiviruses are compared, as elements from distantly related phyla often cluster together without representation from intermediate animal lineages (Fig S4 and Fig 7). Many of these clades tend to have longer branches, suggesting that they are distantly related and diverged a long time ago. For example, within tree position A10 (Fig S4), a clade consisting of elements from Hexanauplia and Annelida are separated by longer branch lengths. The reason for these types of clades, and as to why we do not see elements in species between those distantly related orders, is likely due to the frequent loss of these retrotransposons from the genome. In such cases, distantly related species have seen such retrotransposon copies being active, which are vertically inherited, and diverge along with the genome of the host species, while other species do not see this activity, and those inactive copies would be degraded over time.

The errantiviruses we identified form multiple clades with previously characterized non-env-carrying Ty3/gyspy elements in the pol tree (Fig 7). Additionally, some errantiviruses carry chromodomains, but do not always cluster together with other chromodomain carrying retrotransposons. This pattern could be explained by different lineages of Ty3/gyspy frequently acquiring env from other errantiviruses. This would explain annotated Ty3/gyspy elements forming clades with both HSV/gB- and F-type env-containing errantiviruses. Another explanation could be that widespread horizontal transfer is occurring between distantly related species. As the species diverge from each other, different types of env could be acquired, lost and regained multiple times. However, neither of these explanations account for the congruence seen between pol and the host, at least at the level of animal classes and orders (Fig S4), nor the structural divergence of HSV/gB-type ENV (Fig 5 and S6). The spread of errantiviruses is indicative of multiple lineages of retrotransposon contributing to the current picture of errantiviruses, but the congruence supports their ancient nature. One plausible explanation is that different lineages of Ty3/gyspy within the last common ancestor acquired env, which were inherited and diverged with the host species, infrequently recombining with other elements within the same/similar hosts.

Acquisition of ENV proteins by errantiviruses

It was originally speculated that errantiviruses acquired env from baculoviruses in insects, some of which carry an F-type ENV protein Ld130, potentially via recombination events (Malik et al. 2000; Rohrmann and Karplus 2001). While Ty3/gypsy elements may have obtained the env gene via recombination, it is unlikely that baculoviruses are the source for the env gene found in errantiviruses, as they are far more widespread than insects. The congruence of phylogeny between env and pol indicates that they have likely co-evolved (Fig 4). This, along with the presence of errantiviruses in ancient host orders such as Cnidaria, Ctenophora and Tunicata, suggests that env-carrying Ty3/gypsy elements are highly ancient transposons, rather than the result of a recent recombination event with an exogenous virus. HSV/gB-type env in errantiviruses appears to be equally ancient, as it is similarly found in ancient animal orders (Fig 1). The arrangement of cysteine bridges that stabilize the structure of the ectodomain, are consistent between elements within the same order, even though pol presents itself separately on the tree (Fig 2 and S6). This implies that the env has diverged with the host but likely undergoes recombination events with other elements within the same species. Though, pol genes with HSV/gB- type env mostly cluster within their own clades and are not mixed with pol with F-type env, indicating that recombination events are not happening frequently enough to randomise their spread, and likely occurred anciently. The origins and acquisition, therefore, of either type of env, likely occurred in the last common ancestor of metazoans.

The survival mechanism of errantiviruses in the genome

In Drosophila, errantiviruses occupy the somatic cell niche in the ovaries that surround the germline, and the expression of env allows virus-like particles to infect the germline to create new copies (Song et al. 1994). As an adaptation, Drosophila species have duplicated the germline transposon defence mechanism to target errantiviruses in the somatic cells of the gonad (Brennecke et al. 2007). It is possible that this is more universal, and that env-containing retrotransposons occupy the cell niches surrounding the germline, and that expression of env allows them to continue to infect the germline. For example, murine IAP retrotransposons that carry env genes also remain infectious, like Drosophila errantiviruses (Ribet et al. 2008). However, these additional ORFs and genes in endogenous retroviral elements are constantly subject to evolutionary pressure from their hosts (Wang and Han 2020; Wang et al. 2020). It is therefore significant that the env gene is as highly conserved as we have observed, and that it does not seem to have been subject to frequent loss and acquisition. Though, the mechanisms of transmission via the action of this env gene remain elusive for most of the elements we identified, except for those in Drosophila species. However, the structures formed by env, for both HSV/gB and glycoprotein F, form the same functional structures as those found in exogenous viruses, required for cell entry (Heldwein et al. 2006; Connolly et al. 2011; McLellan et al. 2013). While entry of HSV also requires three other glycoproteins, gD, gH and gL, gB is specifically required for membrane fusion (Johnson et al. 2011; Jambunathan et al. 2021). The mechanism of cell entry, therefore, is unknown for these errantiviruses that carry gB, and necessitates further investigation regarding their expression. Though, given their highly structured configuration, it is likely that they retain membrane fusion capabilities. Additionally, given the widespread nature of these env-containing elements, it may also be possible that somatic transposon defence mechanisms against them are more common.

In summation, the env gene appears to have been anciently acquired by Ty3/gypsy elements, and has been retained by those elements, diverging as their host species evolved. Today, we see a snapshot of errantiviruses widespread throughout metazoan genomes, continuing to create new insertions to survive long term.

Materials and methods

Genome assemblies used in the study

Reference genome sequences were retrieved from NCBI and grouped into the following insect orders and other metazoan phyla: Diptera, Lepidoptera, Coleoptera, Hymenoptera, Hemiptera, Polyneoptera, and Palaeoptera for insects; Arachnida, Branchiopoda, Chilopoda, Collembola, Diplopoda, Hexanauplia, Malacostraca, Merostomata, Ostracoda, and Thecostraca for Arthropoda; Acanthocephala, Annelida, Brachiopoda, Bryozoa, Dicyemida, Mollusca, Nemertea, Onychophora, Orthonectida, Phoronida, Platyhelminthes, Priapulida, Rotifera, and Tardigrada for Protostomia; Cephalochordata, Echinodermata, Hemichordata, and Tunicata for Deuterostomia; Cnidaria, Ctenophora, and Porifera. We obtained lists of reference genomes in each category in May 2023 using the Entrez Direct command line tool. The list of the genome assemblies used can be found in Supplementary Table S4.

Identification of active genomic copies of Ty3/gypsy errantiviruses

We ran a total of five iterations of tBLASTn (ncbi-blast-2.9.0) to search for the genomic copies of Ty3/gypsy errantiviruses, firstly using the baits of known elements, and subsequently using the elements that were identified after each iteration. We chose new baits that were not represented in the phylogenetic tree of the POL (see below for the detail) from the previous iteration and repeated the process until we no longer found new elements. We used GAG, POL and ENV protein sequences of the Dipteran gypsy errantiviruses from the RepBase (https://www.girinst.org/repbase/) for the first tBLAST search. The bait sequences for all tBLASTn iterations can be found at https://github.com/RippeiHayashi/errantiviruses_2025/tree/main. We ran tBLASTn with default parameters (-evalue 10). We applied different criteria to identify ERVs from species outside insects as we did not have prior knowledge of whether or what kind of ERVs are found in those species. We briefly summarise our approach below.

Genomic regions that showed homology to the POL baits with a bit score of greater than 50 or to the GAG and ENV baits with a bit score of greater than 30 were individually pooled. Genomic regions that had hits from multiple baits were merged for their coordinates using bedtools/2.28.0. Regions that were larger than 1500 nucleotides for POL and 300 nucleotides for GAG or ENV were collected. Following, for insect genomes we looked for regions that carry GAG, POL, ENV sequences that are next to each other in this order allowing up to 500 nucleotides of gaps between them as well as excluding regions of smaller than 4000 nucleotides or larger than 10000 nucleotides. For the other genomes we allowed up to 1000 nucleotides of gaps between POL and ENV and collected regions that had both ORFs that are not smaller than 3000 nucleotides or larger than 10000 nucleotides.

We subsequently checked whether the regions contain uninterrupted ORFs by making translated sequences using transeq from EMBOSS-6.6.0 and excluding ORFs with more than five stop codons. We clustered all the identified GAG, POL and ENV from each genome by Clustal Omega (https://www.ebi.ac.uk/Tools/msa/clustalo/) and identified genomic regions that harbour all three ORFs that are multiplied in the genome. We also searched for ORFs regardless of the copy number when there were no multiplied elements in the genome. When genomic copies were abundantly found in multiple closely related species, we only annotated the elements from representative genomes. Genomes that were annotated and the number of predicted ORFs per genome are summarised in Supplementary Table S4.

The above process yielded abundant putative intact copies of ERVs from insect genomes whereas we took additional steps for the other genomes. For Arthropoda genomes outside insects as well as other Protostomia genomes, we extended the genomic regions of predicted GAG-POL to downstream by 3000 nucleotides and manually annotated ENV ORFs using Benchling (https://www.benchling.com/). The identity of GAG ORFs were further confirmed by HHpred (https://toolkit.tuebingen.mpg.de/tools/hhpred). For Cnidaria, Ctenophora and Deuterostomia genomes outside vertebrates we used transeq to annotate putative GAG and ENV ORFs from 5000 nucleotides downstream and 3000 nucleotides upstream to the predicted POL ORFs, respectively.

We took confident 945 POL ORFs from the first tBLASTn search and made a multiple sequence alignment of the conserved regions of the Integrase domain using mafft-7.505 with the option ’--auto’, and inferred the phylogenetic relationships using iqtree-1.6.12 with the option ’-m rtREV+R4 -bb 1000’ (see more details in the section below). We took 152 representative POL from the phylogenetic tree and used the POL and the corresponding GAG and ENV ORFs from the same genomic regions for a second tBLASTn search. We repeated the above steps after combining the results of the two tBLASTn searches. We did not find any ERVs in Nematoda genomes that have GAG, POL and ENV in this order while the previous study reported non-canonical Nemetoda ERVs that have flipped the order of ENV and POL (Sacco et al. 2022). Thus, we searched for regions that have ENV and POL in this order allowing gaps of up to 2000 nucleotides between them. GAG was subsequently predicted from 5000 nucleotides upstream to the predicted ENV-POL. We extended the search for potential GAG-ENV-POL insertions to other non-insect genomes and found similar elements in Bryozoa and Platyhelminthes. We also found ENV-GAG-POL elements in Arachnida genomes (Medioppia subpectinata and Oppia nitens). These elements were recovered through the search of ENV-POL elements because GAG ORF was short. We did not extend our search to find other arrangements of ORFs.

Additionally, we found that the glycoprotein gene present in mosquito BEL/pao retrotransposons (Dezordi et al. 2020) are also found next to gypsy POL genes. The glycoprotein gene resembles the envelope genes from Chuviridae and is structurally similar to glycoprotein B from Herpesviruses. We identified ten mosquito BEL/pao retrotransposons that carry this glycoprotein from RepBase (BEL-199_AA, BEL-279_AA, BEL- 652_AA, BEL-801_AA, BEL-804_AA, BEL-879_AA, BEL-51_AnFu, BEL-56_AnFu, BEL-81_AnFu, and BEL-71_CQ) and used them as baits for a tBLASTn search against Arthropoda genomes. The tBLASTn search identified 114 gypsy elements with homologous glycoprotein genes from Coleoptera, Hymenoptera,

Polyneoptera, and Hexaneuplia species. Combined together, the second iteration of tBLASTn search identified 451 new elements. The phylogenetic analysis of POL from the elements identified in the first and second tBLASTn runs revealed clades that only contained new elements from the second tBLASTn run, suggesting that the first two tBLASTn runs were not comprehensive. Therefore, we took 34 additional baits to cover all the major clades in the phylogenetic tree for a third tBLASTn search, which identified 61 new elements. We repeated tBLASTn runs five times in total; the fourth and fifth runs used 17 and 7 new baits and identified 49 and 6 new elements, respectively. We stopped the tBLASTn search after five iterations because the phylogenetic analysis of POL did not reveal new clades.

Following, we further curated newly identified ERVs. Firstly, we used hmmer 3.3.1 (http://hmmer.org/) to make sure that the ERVs that we identified are classified as Ty3/gypsy. We compared POL ORFs from the newly identified ERVs against the POL ORFs of known retroelements that are available from the Gypsy Database 2.0 (https://gydb.org/index.php/Main_Page). The Gypsy Database 2.0 collects hmm profiles of Ty1/Copia, Bel/Pao, Ty3/gypsy, and Retrovirus POL proteins. The hmmer search classified all the newly identified ERVs into Ty3/gypsy. Secondly, we ran HHpred from hh-suite 3.3.0 (Steinegger et al. 2019) or the online tool of HHpred (https://toolkit.tuebingen.mpg.de/tools/hhpred) to predict GAG, ENV, and POL ORFs using the default search parameters. For the command line HHpred, we first made multiple sequence alignments using hhblits and the Uniclust database UniRef30/2022_02 (https://uniclust.mmseqs.com/). This was followed by hhsearch to predict domain structures using the collection of pdb70 from March 2022 (https://wwwuser.gwdg.de/~compbiol/data/hhsuite/databases/hhsuite_dbs/). We used default settings to run hhblits and hhsearch. ENV and POL ORFs with large truncations were removed after the domain structure prediction. ERVs that predicted all GAG, ENV and POL ORFs were taken for further analyses.

We additionally ran BLASTn using the genomic regions that cover all three ORFs as baits to interrogate the copy number of ERVs in the genome where they are found. The results are summarised in Supplementary Table S1 where we counted the number of BLASTn hits with a more than 98% identity in more than 98% of their length to the bait sequences. A majority of newly identified ERVs (1386 out of 1512 elements) carry ORFs that are found at least twice in the genome, further validating our approach of finding active or recently active insertions.

Identification of putative long terminal repeats and transfer RNA (tRNA) primer binding sequences

Long terminal repeats (LTRs) of ERVs were identified as follows. Firstly, we fetched the genomic coordinates of ERV insertions and extended 3000 nucleotides at either ends from the Start codon of GAG and the Stop codon of ENV ORFs (POL was taken for the GAG-ENV-POL ERVs) for all species categories except for Cnidaria, Ctenophora and Deuterostomia where we extended 5000 nucleotides at either ends from the Start and Stop codons of POL. We retrieved the genomic sequence of the extended ERVs and ran BLASTn against the same sequence to identify repeated sequences. We considered repeated regions of greater than 99% identity and the tBLASTn bit score between 200 and 5000 as putative LTRs. We excluded LTRs whose distance between them is shorter than 5000 nucleotides or longer than 14000 nucleotides. Following, we searched for putative tRNA primer binding sequences (PBS). tRNA sequences were retrieved from genomic annotations from representative genomes in respective species categories except for Ctenophora where we took tRNA sequences from Protostomia, Cnidaria and Deuterostomia because there were no annotated genomes available at the time of the study. Representative genomes that we used to retrieve tRNA sequences are: GCF_000001215.4 and GCF_013141755.1 for Diptera, GCF_014905235.1 and GCF_905147365.1 for Lepidoptera, GCF_000002335.3 and GCF_917563875.1 for Coleoptera, GCF_003254395.2 and GCF_016802725.1 for Hymenoptera, GCF_014356525.2 and GCF_020184175.1 for Hemiptera, GCF_002891405.2 and GCF_021461395.2 for Polyneoptera, GCF_921293095.1 for Palaeoptera, GCF_000671375.1, GCF_020631705.1, GCF_002217175.1, GCF_016086655.3, GCF_024679095.1, GCF_019059575.1 for Arthropoda, GCF_000326865.2, GCF_020536995.1, GCF_000002985.6, GCF_000715545.1 for Protostomia, GCF_000222465.1 and GCF_022113875.1 for Cnidaria, GCF_013122585.1, GCF_000003605.2, GCF_000003815.2, and GCF_000002235.5 for Deuterostomia. The minimum requirements of the priming events are incompletely understood, but the length of PBS for retrotransposons is generally between 8 and 18 nucleotides and the PBS is positioned near the 3’ end of the 5’ LTR (Wilhelm and Wilhelm 2001). We chose the 12 nucleotides at the 3’ end of tRNAs as putative primer sequences and searched for PBS in the 100 nucleotide-window centred at the 3’ end of the 5’ LTR using bowtie-1.2.3 allowing up to two mismatches. We added “CCA” to the 3’ end of the tRNA sequence when it is not present in the genomic sequences. We successfully identified putative PBSs in 1151 out of 1512 ERVs that we identified in this study. We grouped ERVs based on the type of tRNA used for the PBS and summarised the results in Figures S2, S4 and S7 as well as Supplementary Table S1.

Phylogenetic analysis of POL

We obtained multiple sequence alignments of the RT and Integrase domains of POL as follows. For the RT domain, we first took the HHpred alignments of POL against the RT domains of Moloney Murine Leukaemia Virus (Protein Data bank entry 4MH8_A) and Human Immunodeficiency Virus type 1 (Protein Data bank entry 4G1Q_B), spanning regions that correspond to their fingers, palm, thumb and connection subdomains (Das and Georgiadis 2004), namely the amino acid positions of 28 to 426 and 7 to 405, respectively. However, this process caused small truncations at either ends of the domain because of poor alignments at the individual substructures. To overcome this issue and to facilitate sequence alignments, we extended the N- and C-termini by 10 and 15 aa, respectively, after merging the two predicted regions of RT domains. We then ran mafft-7.505 to generate a multiple sequence alignment, from which we manually trimmed the either ends to have intact fingers, palm, thumb and connection subdomains for all POL ORFs. Similarly, we took HHpred alignments of POL against the Integrase domains of Human Spumaretrovirus (Protein Data bank entry 3OYM_A) and Simian T- lymphotropic Virus type 1 (Protein Data bank entry 7OUH_B), spanning regions that correspond to their N- terminal domain (NTD) and catalytic core domain (CCD) (Engelman and Cherepanov 2014), namely the amino acid positions of 54 to 273 and 8 to 197, respectively. We used the predicted Integrase domains with 15 and 30 aa extensions at the N- and C-termini, respectively, for generating a multiple sequence alignment before manually trimming the ends. We inspected the multiple sequence alignments using Mview (https://www.ebi.ac.uk/Tools/msa/mview/) and removed ERVs that contained truncated RT or Integrase domains from the analyses. We included ORFs from 5 Drosophila errantiviruses (DM176, IDEFIX, ZAM, GYPSY, and Gypsy-5_DGri) in the multiple sequence alignments.

We ran iqtree-1.6.12 with the option ’-m rtREV+R4 -bb 1000’ to generate the phylogenetic trees of all Integrase and RT domains, and those from ERVs found in vertebrate and invertebrate genomes, respectively. We then selected representative elements from clades, taking into consideration the relative distances between nodes in the same clade. For example, if the distance between two nodes is less than 0.1, meaning that there is little – no variations between their sequences, only one node in that position is taken. Additionally, we made sure that the representatives captured the diversity of env found within the clade. For example, if a clade contains both F-type and HSV/gB-type elements, at least one for both is taken, regardless of distance. Furthermore, the phylogeny of the hosts was represented as fully as possible. For example, if a clade has elements from Diptera and Coleoptera, at least one from both is taken regardless of distance. Once representatives were collected, the tree was then accordingly pruned. It should be noted that the final pruned tree of representatives was only used as a figure aid for clarity (Fig 1-4), and all analyses were done on the full Integrase and RT phylogenetic trees. We used a similar approach as detailed to select representatives for structural analysis of env ectodomains, as well as the pol bridge region (Table S2). If any outliers were found, remaining elements from the same clade were also analysed, to check whether structural inconsistencies were consistent across the clade.

For congruence plots in figures 4 and S2, we grouped elements based on which node they sit in the tree, ensuring that the node is well supported by bootstrap values greater than 95. We used the networkD3 package in R to generate Sankey plots to visualise the representation of elements from one tree in another. The node definitions are summarized in Table S3.

Further, we took 86 representatives of the F-type ENV based on their positions in the corresponding POL phylogenetic tree for the analysis. We identified the signal peptides and transmembrane regions by DeepTMHMM (https://dtu.biolib.com/DeepTMHMM) and obtained the ectodomains that are in between them. We subsequently ran mafft-7.505 to generate a multiple sequence alignment.

Structure prediction of ENV proteins, the RNase H and chromodomains using Alphafold 2

We took representatives of the F-type and HSV/gB-type ENV proteins and the region of POL between RT and Integrase, that we termed the bridge region, using branch definitions from the phylogenetic trees of pol RT. We also took the C-terminal half of POL, including the Integrase domain, from representative errantiviruses, investigated in Figure 7. We predicted their structures when chromodomains were detected by hhpred (see above). We used the colabfold wrap v1.4.0 of Alphafold 2 (https://github.com/YoshitakaMo/localcolabfold/releases/tag/v1.4.0), and used MMSeqs2 for the database search and clustering of protein sequences before performing the structural prediction (Mirdita et al. 2022). We used the whole ectodomains (see above) of the F-type and HSV/gB-type ENV proteins and predicted the trimer structures using the multimer mode of Alphafold 2 (AlphaFold2-multimer-v2). We then visualised and characterised visible structures and disulfide bonds or cysteine bridges using Maestro (Schrödinger Release 2024-3) and grouped them by common substructures (Table S2). Structures of all predicted chromodomains in POL were confirmed by Alphafold.

eLife Assessment

This important study provides convincing evidence that envelope-carrying Ty3/gypsy retrotransposons (errantiviruses) are ancient, widespread, and actively expanding across nearly all major animal phyla. Using comprehensive phylogenetic and AlphaFold2-based structural analyses, the authors show that these elements independently acquired membrane fusion proteins early in metazoan evolution, likely predating the bilaterian-non-bilaterian split. While some aspects could be more clearly contextualized and explained better, the work offers insights into the deep evolutionary roots of retroelement-envelope associations and the origins of retroviruses.

Reviewer #1 (Public review):

Summary:

This manuscript provides a comprehensive systematic analysis of envelope-containing Ty3/gypsy retrotransposons (errantiviruses) across metazoan genomes, including both invertebrates and ancient animal lineages. Using iterative tBLASTn mining of over 1,900 genomes, the authors catalog 1,512 intact retrotransposons with uninterrupted gag, pol, and env open reading frames. They show that these elements are widespread-present in most metazoan phyla, including cnidarians, ctenophores, and tunicates-with active proliferation indicated by their multicopy status. Phylogenetic analyses distinguish "ancient" and "insect" errantivirus clades, while structural characterization (including AlphaFold2 modeling) reveals two major env types: paramyxovirus F-like and herpesvirus gB-like proteins. Although bot envelope types were identified in previous analyses two decades ago, the evolutionary provenance of these envelope genes was almost rudimentary and anecdotal (I can say this because I authored one of these studies). The results in the present study support an ancient origin for env acquisition in metazoan Ty3/gypsy elements, with subsequent vertical inheritance and limited recombination between env and pol domains. The paper also proposes an expanded definition of 'errantivirus' for env-carrying Ty3/gypsy elements outside Drosophila.

Strengths:

(1) Comprehensive Genomic Survey:

The breadth of the genome search across non-model metazoan phyla yields an impressive dataset covering evolutionary breadth, with clear documentation of search iterations and validation criteria for intact elements.

(2) Robust Phylogenetic Inference:

The use of maximum likelihood trees on both pol and env domains, with thorough congruence analysis, convincingly separates ancient from lineage-specific elements and demonstrates co-evolution of env and pol within clades.

(3) Structural Insights:

AlphaFold2-based predictions provide high-confidence structural evidence that both env types have retained fusion-competent architectures, supporting the hypothesis of preserved functional potential.

(4) Novelty and Scope:

The study challenges previous assumptions of insect-centric or recent env acquisition and makes a compelling case for a Pre-Cambrian origin, significantly advancing our understanding of animal retroelement diversity and evolution. THIS IS A MAJOR ADVANCE.

(5) Data Transparency:

I appreciate that all data, code, and predicted structures are made openly available, facilitating reproducibility and future comparative analyses.

Major Weaknesses

(1) Functional Evidence Gaps:

The work rests largely on sequence and structure prediction. No direct expression or experimental validation of envelope gene function or infectivity outside Drosophila is attempted, which would be valuable to corroborate the inferred roles of these glycoproteins in non-insect lineages. At least for some of these species, there are RNA-seq datasets that could be leveraged.

(2) Horizontal Transfer vs. Loss Hypotheses:

The discussion argues primarily for vertical inheritance, but the somewhat sporadic phylogenetic distributions and long-branch effects suggest that loss and possibly rare horizontal events may contribute more than acknowledged. Explicit quantitative tests for horizontal transfer, or reconciliation analyses, would strengthen this conclusion. It's also worth pointing out that, unlike retrotransposons that can be found in genomes, any potential related viral envelopes must, by definition, have a spottier distribution due to sampling. I don't think this challenges any of the conclusions, but it must be acknowledged as something that could affect the strength of this conclusion

(3) Limited Taxon Sampling for Certain Phyla:

Despite the impressive breadth, some ancient lineages (e.g., Porifera, Echinodermata) are negative, but the manuscript does not fully explore whether this reflects real biological absence, assembly quality, or insufficient sampling. A more systematic treatment of negative findings would clarify claims of ubiquity. However, I also believe this falls beyond the scope of this study.

(4) Mechanistic Ambiguity:

The proposed model that env-containing elements exploit ovarian somatic niches is plausible but extrapolated from Drosophila data; for most taxa, actual tissue specificity, lifecycle, or host interaction mechanisms remain speculative and, to me, a bit unreasonable.

Minor Weaknesses:

(1) Terminology and Nomenclature:

The paper introduces and then generalizes the term "errantivirus" to non-insect elements. While this is logical, it may confuse readers familiar with the established, Drosophila-centric definition if not more explicitly clarified throughout. I also worry about changes being made without any input from the ICTV nomenclature committee, which just went through a thorough reclassification. Nevertheless, change is expected, and calling them all errantiviruses is entirely reasonable.

(2) Figures and Supplementary Data Navigation:

Some key phylogenies and domain alignments are found only in supplementary figures, occasionally hindering readability for non-expert audiences. Selected main-text inclusion of representative trees would benefit accessibility.

(3) ORF Integrity Thresholds:

The cutoff choices for defining "intact" elements (e.g., numbers/placement of stop codons, length ranges) are reasonable but only lightly justified. More rationale or sensitivity analysis would improve confidence in the inclusion criteria. For example, how did changing these criteria change the number of intact elements?

(4) Minor Typos/Formatting:

The paper contains sporadic typographical errors and formatting glitches (e.g., misaligned figure labels, unrendered symbols) that should be addressed.

Reviewer #2 (Public review):

Summary:

The authors first surveyed metazoan genomes to identify homologs of Drosophila errantiviruses and classified them into two groups, "insect" and "ancient" elements, supporting the hypothesis of an early evolutionary origin for these retrotransposons. They subsequently identified two distinct types of envelope proteins, one resembling the glycoprotein F of paramyxoviruses and the other akin to the glycoprotein B of herpesviruses. Despite differences in their primary amino acid sequences, these proteins display notable structural similarity in their predicted domain architectures. The congruence between the phylogenies of the envelope and pol genes further supports the ancient origin of the envelope genes, challenging earlier hypotheses that proposed recent recombination events with baculoviruses. Additional analysis of the Pol "bridge region" corroborated the divergence among these elements, consistent with a pattern of limited cross-species recombination. Finally, by comparing these elements with non-envelope-containing Gypsy retrotransposons, the authors concluded that errantiviruses originated from multiple elements independently.

Strengths:

The conclusions of this study are based on a comprehensive collection of errantiviruses identified across a wide range of metazoan genomes. These findings are further supported by multiple lines of evidence, including phylogenetic congruence and the diverse evolutionary origins of envelope genes. AlphaFold2-assisted protein domain structure analyses also provided key insights into the characterization of these elements. Together, these results present a compelling case that errantiviruses arose independently through multiple evolutionary events, extending well beyond previous hypotheses.

Weaknesses:

It would be beneficial to emphasize in the Abstract the potential impact of this work by more clearly articulating the current knowledge gap in the field. While the second paragraph of the Introduction briefly touches on this point, highlighting the broader significance in the Abstract would better capture readers' interest. Additionally, some methodological choices would benefit from clearer justification and explanation. For instance, in Figure 6, the selection of the bridge region/RNase H domain is not explicitly explained, leaving the rationale for its choice unclear. As a minor point, some figure labels and texts are too small and difficult to read, and improving their legibility would enhance overall clarity.

Reviewer #3 (Public review):

Summary and Significance:

In this work, Cary and Hayashi address the important question of when, in evolution, certain mobile genetic elements (Ty3/gypsy-like non-LTR retrotransposons) associated with certain membrane fusion proteins (viral glycoprotein F or B-like proteins), which could allow these mobile genetic elements to be transferred between individual cells of a given host. It is debated in the literature whether the acquisition of membrane fusion proteins by non-LTR retrotransposons is a rather recent phenomenon that separately occurred in the ancestors of certain host species or whether the association with membrane fusion proteins is a much more ancient one, pre-dating the Cambrian explosion. Obviously, this question also touches upon the origin of the retroviruses, which can spread between individuals of a given host but seem restricted to vertebrates. Based on convincing data, Cary and Hayashi argue that an ancient association of non-LTR retrotransposons with membrane fusion proteins is most probable.

Strengths:

The authors take the smart approach to systematically retrieve apparently complete, intact, and recently functional Ty3/gypsy-like non-LTR retrotransposons that, next to their characteristic gag and pol genes, additionally carry sequences that are homologous to viral glycoprotein F (env-F) or viral glycoprotein B (env-B). They then construct and compare phylogenetic trees of the host species and individual encoded proteins and protein domains, where 3D-structure calculations and other features explain and corroborate the clustering within the phylogenetic trees. Congruence of phylogenetic trees and correlation of structural features is then taken as evidence for an infrequent recombination and a long-term co-evolution of the reverse transcriptase (encoded by the pol gene) and its respective putative membrane fusion gene (encoded by env-F or env-B). Importantly, the env-F and env-B containing retrotransposons do not form a monophyletic group among the Ty3/gypsy-like non-LTR retrotransposons, but are scattered throughout, supporting the idea of an originally ancient association followed by a random loss of env-F/env-B in individual branches of the tree (and rather rare re-associations via more recent recombinations).

Overall, this is valuable, stimulating, and important work of general and fundamental interest, but still also somewhat incompletely explored, imprecisely explained, and insufficiently put into context for a more general audience.

Weaknesses:

Some points that might be considered and clarified:

(1) Imprecise explanations, terms, and definitions:

It might help to add a 'definitions box' or similar to precisely explain how the authors decided to use certain terms in this manuscript, and then use these terms consistently and with precision.

a) In particular, these are terms such as 'vertebrate retrovirus' vs 'retrovirus' vs 'endogenized retrovirus' vs 'endogenous retrovirus' vs 'non-LTR retrotransposon' and 'Ty3/gypsi-like retrotransposon' vs 'Ty3/gypsy retrotransposon' vs 'errantivirus'.

b) The comment also applies to the term 'env' used for both 'env-F' and 'env-B', where often it remains unclear which of the two protein types the authors refer to. This is confusing, particularly in the methods, where the search for the respective homologs is described.

c) Other examples are the use of the entire pol gene vs. pol-RT for the definition of the Ty3/gypsy clade and for the generation of phylogenetic trees (Methods and Figure S1), and the names for various portions of pol that appear without prior definition or explanation (e.g., 'pro' in Figure 1A, 'bridge' in Figure S1C, 'the chromodomain' in the text and Figure 7).

d) It is unclear from the main text which portions of pol were chosen to define pol-RT and why. The methods name the 'palm-and-fingers', 'thumb', and 'connections' domains to define RT. In the main text, the 'connection' domain is called 'tether' and is instead defined as part of the 'bridge' region following RT, which is not part of RT.

(2) Insufficient broader context:

a) The introduction does not state what defines Ty3/gypsy non-LTR retrotransposons as compared to their closest relatives (Ty1/copia retrotransposons, BEL/pao retrotransposons, vertebrate retroviruses). This makes it difficult to judge the significance and generality of the findings.

b) The various known compositions of Ty3/gypsi-like retrotransposons are not mentioned and explained in the introduction (open reading frames, (poly-)proteins and protein domains, and their variable arrangement, enzymatic activities, and putative functions), and the distribution of Ty3/gypsi-like retrotransposons among eukaryotes remains unclear. The introduction does not mention that Ty3/gypsi-like retrotransposons apparently are absent from vertebrates, and Figure 7 is not very clear about whether or not it includes sequences from plants ('Chromoviridae').

c) The known association of Ty3/gypsi-like retrotransposons from different metazoan phyla with putative membrane fusion proteins (env-like) genes is mentioned in the introduction, but literature information, whether such associations also occur in the context of other retrotransposons (e.g., Ty1/ copia or BEL/pao), is not provided. The abstract is somewhat misleading in this respect. Finally, the different known types of env-like genes are not mentioned and explained as part of the introduction ('env-f', 'env-B', 'retroviral env', others?)

d) Some key references and reviews might be added:

- Pelisson, A. et al. (1994) https://www.embopress.org/doi/abs/10.1002/j.1460-2075.1994.tb06760.x

(next to Song et al. (1994), for the identification of env in Ty3/gypsy)

- Boeke, J.D. et al. (1999)

In Virus Taxonomy: ICTV VIIth report. (ed. F.A. Murphy),. Springer-Verlag, New York.

(cited by Malik et al. (2000) - for the definition and first use of the term 'errantivirus')

- Eickbush, T.H. and Jamburuthugoda, V.K. (2008) https://doi.org/10.1016/j.virusres.2007.12.010

(on the classification of retrotransposons and their env-like genes)

- Hayward, A. (2017) https://doi.org/10.1016/j.coviro.2017.06.006

(on scenarios of env acquisition)

(3) Incomplete analysis:

a) Mobile genetic elements are sometimes difficult to assemble correctly from short-read sequencing data. Did the authors confirm some of their newly identified elements by e.g., PCR analysis or re-identification in long-read sequencing data?

b) The authors mention somewhat on the side that there are Ty3/gypsy elements with a different arrangement (gag-env-pol instead of gag-pol-env). Why was this important feature apparently not used and correlated in the analysis? How does it map on the RT phylogenetic tree? Which type of env is found with either arrangement? Is there evidence for a loss of env also in the case of gag-env-pol elements?

c) Sankey plots are insufficiently explained. How would inconsistencies between trees (recombinations) show up here? Why is there no Sankey plot for the analysis of env-B in Figure 5?

d) Why are there no trees generated for env-F and env-B like proteins, including closely related homologous sequences that do NOT come from Ty3/gypsy retrotransposons (e.g., from the eukaryotic hosts, from other types of retrotransposons (Ty1/copia or BEL/pao), from viruses such as Herpesvirus and Baculovirus)? It would be informative whether the sequences from Ty3/gypsy cluster together in this case.

e) Did the authors identify any other env-like ORFs (apart from env-F and env-B) among Ty3/gypsy retrotransposons? Did they identify other, non-env-like ORFs that might help in the analysis? It is not quite clear from the methods if the searches for env-F and env-B - containing Ty3/gypsy elements were done separately and consecutively or somehow combined (the authors generally use 'env', and it is not clear which type of protein this refers to).

f) Why was the gag protein apparently not used to support the analysis? Are there different, unrelated types of gag among non-LTR retrotransposons? Does gag follow or break the pattern of co-evolution between RT and env-F/env-B?

g) Data availability. The link given in the paper does not seem to work (https://github.com/RippeiHayashi/errantiviruses_2025/tree/main). It would be useful for the community to have the sequences of the newly identified Ty3/gypsy retrotransposons listed readily available (not just genome coordinates as in table S1), together with the respective annotations of ORFs and features.

Author response:

The following is the authors’ response to the original reviews.

In the revised manuscript, we have clarified several points that were raised by the reviewers. First, we now state more explicitly that the presence of intact env-containing Ty3/gypsy retrotransposons does not by itself demonstrate their mechanism of transmission, tissue specificity, infectivity, or current activity. We have therefore revised the wording throughout the manuscript to distinguish intact element structure and multicopy genomic expansion from experimentally demonstrated activity.

Second, we performed targeted host-taxonomy concordance analyses on selected clades of the POL RT tree. These analyses do not exclude local horizontal transfer, particularly between closely related hosts, but they show that horizontal transfer alone is insufficient to explain the broader host-taxonomic structure observed across the dataset.

Third, we incorporated representative viral and retroelement-associated fusogen proteins into our F-type ENV phylogenetic analysis and HSV/gB-type ENV structural comparison. These additions place the ENV proteins associated with Ty3/gypsy elements in a broader evolutionary context and strengthen the conclusion that these ENV associations are deeply diverged rather than recent derivatives of a single sampled viral lineage.

Fourth, we added two each of entirely new Supplementary figures (S5 and S7) and Tables (S2 and S3) and substantially modified now Supplementary figure S8. Other figures have also been modified only to increase readability. The four tables from the original manuscript have not been modified although their numbering has changed.

We believe that the revised manuscript is substantially improved in clarity, terminology and interpretive precision, while retaining the central conclusion that the association between env-like genes and Ty3/gypsy retrotransposons is ancient in metazoan evolution. Sincerely,

Public Reviews:

Reviewer #1 (Public review):

Summary:

This manuscript provides a comprehensive systematic analysis of envelope-containing Ty3/gypsy retrotransposons (errantiviruses) across metazoan genomes, including both invertebrates and ancient animal lineages. Using iterative tBLASTn mining of over 1,900 genomes, the authors catalog 1,512 intact retrotransposons with uninterrupted gag, pol, and env open reading frames. They show that these elements are widespread present in most metazoan phyla, including cnidarians, ctenophores, and tunicates-with active proliferation indicated by their multicopy status. Phylogenetic analyses distinguish "ancient" and "insect" errantivirus clades, while structural characterization (including AlphaFold2 modeling) reveals two major env types: paramyxovirus F-like and herpesvirus gB-like proteins. Although bot envelope types were identified in previous analyses two decades ago, the evolutionary provenance of these envelope genes was almost rudimentary and anecdotal (I can say this because I authored one of these studies). The results in the present study support an ancient origin for env acquisition in metazoan Ty3/gypsy elements, with subsequent vertical inheritance and limited recombination between env and pol domains. The paper also proposes an expanded definition of 'errantivirus' for env-carrying Ty3/gypsy elements outside Drosophila.

Strengths:

(1) Comprehensive Genomic Survey:

The breadth of the genome search across non-model metazoan phyla yields an impressive dataset covering evolutionary breadth, with clear documentation of search iterations and validation criteria for intact elements.

(2) Robust Phylogenetic Inference:

The use of maximum likelihood trees on both pol and env domains, with thorough congruence analysis, convincingly separates ancient from lineage-specific elements and demonstrates co-evolution of env and pol within clades.

(3) Structural Insights:

AlphaFold2-based predictions provide high-confidence structural evidence that both env types have retained fusion-competent architectures, supporting the hypothesis of preserved functional potential.

(4) Novelty and Scope:

The study challenges previous assumptions of insect-centric or recent env acquisition and makes a compelling case for a Pre-Cambrian origin, significantly advancing our understanding of animal retroelement diversity and evolution. THIS IS A MAJOR ADVANCE.

(5) Data Transparency:

I appreciate that all data, code, and predicted structures are made openly available, facilitating reproducibility and future comparative analyses.

Major Weaknesses

(1) Functional Evidence Gaps:

The work rests largely on sequence and structure prediction. No direct expression or experimental validation of envelope gene function or infectivity outside Drosophila is attempted, which would be valuable to corroborate the inferred roles of these glycoproteins in non-insect lineages. At least for some of these species, there are RNA-seq datasets that could be leveraged.

We added a sentence in the discussion, subsection “The survival mechanism of errantiviruses in the genome”, citing our recent work now published (PMID: 41922845), explaining that the defence mechanism against errantiviruses appears to be conserved in insects beyond Drosophila, indirectly suggesting that their biology dependent on the presence of env—may be more universal.

(2) Horizontal Transfer vs. Loss Hypotheses:

The discussion argues primarily for vertical inheritance, but the somewhat sporadic phylogenetic distributions and long-branch effects suggest that loss and possibly rare horizontal events may contribute more than acknowledged. Explicit quantitative tests for horizontal transfer, or reconciliation analyses, would strengthen this conclusion. It's also worth pointing out that, unlike retrotransposons that can be found in genomes, any potential related viral envelopes must, by definition, have a spottier distribution due to sampling. I don't think this challenges any of the conclusions, but it must be acknowledged as something that could affect the strength of this conclusion

We have added a targeted host-taxonomy concordance analysis for two well-sampled POL extended RT/connection subclades: an Annelida-associated clade from tree position A6 and a Lepidoptera-associated clade from tree position I1 (new Fig S5). Rather than attempting to infer exact numbers of duplication, loss and horizontal transfer events, which is difficult across highly expanded and unevenly sampled transposon families, we tested whether host-taxonomic labels were more clustered on the observed POL extended RT/connection topology than expected by chance. In the Annelida clade, highly supported small subclades showed strong host-family and host species concordance under host-label permutation tests. The Lepidoptera clade showed a more mixed pattern, but still contained several highly supported subclades enriched for related host groups at the superfamily or broader taxonomic level. These results do not exclude rare horizontal transfer, particularly between closely related hosts, but support the conclusion that the observed POL extended RT/connection trees retain significant host-taxonomic structure and are not consistent with frequent broad horizontal transfer between distantly related animal groups. We have added a paragraph in the Results section “Multiple intact elements of env-carrying Ty3/gypsy retrotransposons are found widespread across metazoan species” describing these observations, and also revised the Discussion to more explicitly acknowledge the possibilities of lineage-specific loss and the horizontal transfer.

(3) Limited Taxon Sampling for Certain Phyla:

Despite the impressive breadth, some ancient lineages (e.g., Porifera, Echinodermata) are negative, but the manuscript does not fully explore whether this reflects real biological absence, assembly quality, or insufficient sampling. A more systematic treatment of negative findings would clarify claims of ubiquity. However, I also believe this falls beyond the scope of this study.

In the revised manuscript, we have added a targeted analysis of two representative genomes from each phylum. Although we did not detect full-length GAG-POL-ENV elements in these genomes, we recovered multiple full-length, multicopy GAG-POL Ty3/gypsy elements from all four genomes, many of which were flanked by predicted LTR sequences and associated with putative tRNA primer-binding sites. This suggests that the apparent absence of env-carrying elements in these representative Porifera and Echinodermata genomes is unlikely to be due simply to poor assembly quality or a general inability to recover intact Ty3/gypsy-like retrotransposons. We have added these data as the new Supplementary table S3 and revised the Results section “Multiple intact elements of env-carrying Ty3/gypsy retrotransposons are found widespread across metazoan species” to clarify that absence in these phyla may reflect true biological absence, lineage-specific loss, or incomplete taxon sampling.

(4) Mechanistic Ambiguity:

The proposed model that env-containing elements exploit ovarian somatic niches is plausible but extrapolated from Drosophila data; for most taxa, actual tissue specificity, lifecycle, or host interaction mechanisms remain speculative and, to me, a bit unreasonable.

We stressed in the Discussion section “The survival mechanism of errantiviruses in the genome” that the mere presence of env gene does not imply the mechanism of transmission of retrotransposons.

Minor Weaknesses:

(1) Terminology and Nomenclature:

The paper introduces and then generalizes the term "errantivirus" to non-insect elements. While this is logical, it may confuse readers familiar with the established, Drosophila-centric definition if not more explicitly clarified throughout. I also worry about changes being made without any input from the ICTV nomenclature committee, which just went through a thorough reclassification. Nevertheless, change is expected, and calling them all errantiviruses is entirely reasonable.

We have revised the Results section and discussion where we introduced the term "errantivirus" to clarify that we use "errantivirus" operationally to refer to env-containing Ty3/gypsy retrotransposons identified in this study, rather than as a formal taxonomic proposal. We also now state explicitly that bona fide infectivity and amplification through the Drosophila-like ovarian somatic-cell route have not been experimentally established for most non-Drosophila elements. Our use of the term is therefore intended to distinguish env-containing Ty3/gypsy elements from related non-env containing Ty3/gypsy retrotransposons, while acknowledging that their biology outside Drosophila remains to be determined.

(2) Figures and Supplementary Data Navigation:

Some key phylogenies and domain alignments are found only in supplementary figures, occasionally hindering readability for non-expert audiences. Selected main-text inclusion of representative trees would benefit accessibility.

We agree that clearer navigation between the main text and supplementary figures would improve readability. Although we considered moving selected supplementary phylogenies and alignments into the main figures, the main figures are already data-dense and are intended to provide representative summaries across many host groups and ENV types. We therefore retained the detailed trees and alignments as supplementary figures, where they can be shown at readable scale, but revised the manuscript to improve navigation. Specifically, we added signposting sentences in the Results where supplementary figures are mentioned, expanded the relevant figure legends, and clarified how each supplementary tree or alignment supports the corresponding main-text conclusion.

(3) ORF Integrity Thresholds:

The cutoff choices for defining "intact" elements (e.g., numbers/placement of stop codons, length ranges) are reasonable but only lightly justified. More rationale or sensitivity analysis would improve confidence in the inclusion criteria. For example, how did changing these criteria change the number of intact elements?

We agree with the reviewer that the rationale for the ORF integrity thresholds should be stated more clearly. We have revised the Methods section "Identification of intact genomic copies of Ty3/gypsy errantiviruses" to clarify that the initial length, gap and stop-codon thresholds were deliberately permissive screening criteria, designed to avoid excluding divergent or non-canonical elements at the discovery stage. These initial filters were not used alone to define the final “intact” set. Candidate elements were subsequently subjected to multiple additional curation steps, including confirmation of Ty3/gypsy POL identity, recovery of full-length RT and Integrase domains within continuous ORFs, HHpred-based domain annotation of GAG, POL and ENV, and removal of elements with large domain truncations. Thus, the final set of intact elements is substantially more refined than would be implied by the initial stop codon or length thresholds alone.

A full sensitivity analysis varying each threshold across the entire iterative discovery and manual-curation pipeline would be difficult to interpret, because changing early permissive filters would alter the candidate pool that then undergoes downstream structural and phylogenetic validation. Instead, we have clarified in the Methods that the early thresholds were intended as inclusive prefilters, whereas final inclusion required intact domain architecture and phylogenetic/domain support.

(4) Minor Typos/Formatting:

The paper contains sporadic typographical errors and formatting glitches (e.g., misaligned figure labels, unrendered symbols) that should be addressed.

We now fixed these issues in the revised manuscript.

Reviewer #2 (Public review):

Summary:

The authors first surveyed metazoan genomes to identify homologs of Drosophila errantiviruses and classified them into two groups, "insect" and "ancient" elements, supporting the hypothesis of an early evolutionary origin for these retrotransposons. They subsequently identified two distinct types of envelope proteins, one resembling the glycoprotein F of paramyxoviruses and the other akin to the glycoprotein B of herpesviruses. Despite differences in their primary amino acid sequences, these proteins display notable structural similarity in their predicted domain architectures. The congruence between the phylogenies of the envelope and pol genes further supports the ancient origin of the envelope genes, challenging earlier hypotheses that proposed recent recombination events with baculoviruses. Additional analysis of the Pol "bridge region" corroborated the divergence among these elements, consistent with a pattern of limited cross-species recombination. Finally, by comparing these elements with non-envelope-containing Gypsy retrotransposons, the authors concluded that errantiviruses originated from multiple elements independently.

Strengths:

The conclusions of this study are based on a comprehensive collection of errantiviruses identified across a wide range of metazoan genomes. These findings are further supported by multiple lines of evidence, including phylogenetic congruence and the diverse evolutionary origins of envelope genes. AlphaFold2-assisted protein domain structure analyses also provided key insights into the characterization of these elements. Together, these results present a compelling case that errantiviruses arose independently through multiple evolutionary events, extending well beyond previous hypotheses.

Weaknesses:

It would be beneficial to emphasize in the Abstract the potential impact of this work by more clearly articulating the current knowledge gap in the field. While the second paragraph of the Introduction briefly touches on this point, highlighting the broader significance in the Abstract would better capture readers' interest. Additionally, some methodological choices would benefit from clearer justification and explanation. For instance, in Figure 6, the selection of the bridge region/RNase H domain is not explicitly explained, leaving the rationale for its choice unclear. As a minor point, some figure labels and texts are too small and difficult to read, and improving their legibility would enhance overall clarity.

We have revised the Abstract to more clearly state the knowledge gap addressed by this study: although env-containing Ty3/gypsy elements were known from Drosophila and sporadically reported in other animals, whether their association with env-like fusogen genes reflected recent, lineage-specific acquisitions or a much deeper evolutionary relationship remained unclear. We now highlight this broader significance in the Abstract and frame our results as evidence that env-containing Ty3/gypsy elements represent deeply diverged, genome-resident retroelements rather than a recent insect-specific phenomenon.

We have also revised the Results, Methods and Figure 6 legend to explain why the

RNase H-containing bridge region was analysed. Specifically, we now distinguish the Pol extended RT/connection region used for phylogenetic analysis from the RNase H-containing bridge region analysed structurally in Figure 6. We define the bridge region as the canonical RNase H domain together with the C-terminal region between RNase H and Integrase, and explain that this region was selected because RNase H-related and adjacent RNase H-like domains vary among LTR retroelement lineages. The bridge region architecture therefore provides an independent structural feature for comparing the “insect errantivirus” and “ancient errantivirus” groups.

Finally, we have revised the figures and figure legends to improve readability. In particular, we enlarged labels where possible, clarified figure annotations, corrected cross-references between main and supplementary figures, and added signposting sentences in the Results so that readers can more easily connect the main conclusions to the supporting supplementary trees and alignments.

Reviewer #3 (Public review):

Summary and Significance:

In this work, Cary and Hayashi address the important question of when, in evolution, certain mobile genetic elements (Ty3/gypsy-like non-LTR retrotransposons) associated with certain membrane fusion proteins (viral glycoprotein F or B-like proteins), which could allow these mobile genetic elements to be transferred between individual cells of a given host. It is debated in the literature whether the acquisition of membrane fusion proteins by non-LTR retrotransposons is a rather recent phenomenon that separately occurred in the ancestors of certain host species or whether the association with membrane fusion proteins is a much more ancient one, pre-dating the Cambrian explosion. Obviously, this question also touches upon the origin of the retroviruses, which can spread between individuals of a given host but seem restricted to vertebrates. Based on convincing data, Cary and Hayashi argue that an ancient association of non-LTR retrotransposons with membrane fusion proteins is most probable.

Strengths:

The authors take the smart approach to systematically retrieve apparently complete, intact, and recently functional Ty3/gypsy-like non-LTR retrotransposons that, next to their characteristic gag and pol genes, additionally carry sequences that are homologous to viral glycoprotein F (env-F) or viral glycoprotein B (env-B). They then construct and compare phylogenetic trees of the host species and individual encoded proteins and protein domains, where 3D-structure calculations and other features explain and corroborate the clustering within the phylogenetic trees. Congruence of phylogenetic trees and correlation of structural features is then taken as evidence for an infrequent recombination and a long-term co-evolution of the reverse transcriptase (encoded by the pol gene) and its respective putative membrane fusion gene (encoded by env-F or env-B). Importantly, the env-F and env-B containing retrotransposons do not form a monophyletic group among the Ty3/gypsy-like non-LTR retrotransposons, but are scattered throughout, supporting the idea of an originally ancient association followed by a random loss of env-F/env-B in individual branches of the tree (and rather rare re-associations via more recent recombinations).

Overall, this is valuable, stimulating, and important work of general and fundamental interest, but still also somewhat incompletely explored, imprecisely explained, and insufficiently put into context for a more general audience.

Weaknesses:

Some points that might be considered and clarified:

(1) Imprecise explanations, terms, and definitions:

It might help to add a 'definitions box' or similar to precisely explain how the authors decided to use certain terms in this manuscript, and then use these terms consistently and with precision.

(a) In particular, these are terms such as 'vertebrate retrovirus' vs 'retrovirus' vs 'endogenized retrovirus' vs 'endogenous retrovirus' vs 'non-LTR retrotransposon' and 'Ty3/gypsi-like retrotransposon' vs 'Ty3/gypsy retrotransposon' vs 'errantivirus'.

We agree with the reviewer. We inserted a paragraph at the end of the first Results section, explaining how we define endogenous retroviruses (ERVs), Ty3/gypsy retrotransposons and errantiviruses.

(b) The comment also applies to the term 'env' used for both 'env-F' and 'env-B', where often it remains unclear which of the two protein types the authors refer to. This is confusing, particularly in the methods, where the search for the respective homologs is described.

We revised the manuscript and now used F-type env/ENV and HSV/gB-type env/ENV throughout the text. We also modified the method section where we explained the tBlastn search to clarify which ENV proteins were used initially for the search and how we classified them in later analyses.

(c) Other examples are the use of the entire pol gene vs. pol-RT for the definition of the Ty3/gypsy clade and for the generation of phylogenetic trees (Methods and Figure S1), and the names for various portions of pol that appear without prior definition or explanation (e.g., 'pro' in Figure 1A, 'bridge' in Figure S1C, 'the chromodomain' in the text and Figure 7).

We revised the manuscript and explained ‘pro’, ‘bridge’ and ‘the chromodomain’ in the Results section or figure legends when they first appear. Please refer to other sections of the response for pol-RT definition.

(d) It is unclear from the main text which portions of pol were chosen to define pol-RT and why. The methods name the 'palm-and-fingers', 'thumb', and 'connections' domains to define RT. In the main text, the 'connection' domain is called 'tether' and is instead defined as part of the 'bridge' region following RT, which is not part of RT.

We agree that our previous terminology around Pol domains was imprecise and could confuse readers. We have revised the manuscript to distinguish the region used for phylogenetic analysis from the region analysed structurally in Figure 6. The phylogenetic analysis used an extended RT/connection region, comprising the RT polymerase core together with the downstream connection subdomain. This connection subdomain is treated as part of retroviral RT in structural studies, but corresponds to a partial RNase H-like fold and has been interpreted evolutionarily as a degenerated RNase H-like tether domain. It is therefore broader than the RT polymerase core alone, but it is not the complete canonical RNase H domain.

We now define the Figure 6 “bridge region” separately as the region spanning the canonical RNase H domain and the C-terminal region between RNase H and Integrase. Figure 6 shows that the invertebrate errantiviruses analysed retain an intact canonical RNase H domain immediately downstream of the extended RT/connection region, but differ in the additional downstream RNase H-like or mini-domain structures before Integrase. We have revised the Results, Methods and figure legends accordingly. We also acknowledge that a phylogeny based strictly on the RT polymerase core alone could differ in some local branch relationships, but the major conclusions are supported independently by the Integrase tree, host-taxonomic structure, ENV-type distribution, Pol bridge-region architecture and ENV structural features.

(2) Insufficient broader context:

(a) The introduction does not state what defines Ty3/gypsy non-LTR retrotransposons as compared to their closest relatives (Ty1/copia retrotransposons, BEL/pao retrotransposons, vertebrate retroviruses). This makes it difficult to judge the significance and generality of the findings.

(b) The various known compositions of Ty3/gypsi-like retrotransposons are not mentioned and explained in the introduction (open reading frames, (poly-)proteins and protein domains, and their variable arrangement, enzymatic activities, and putative functions), and the distribution of Ty3/gypsi-like retrotransposons among eukaryotes remains unclear. The introduction does not mention that Ty3/gypsi-like retrotransposons apparently are absent from vertebrates, and Figure 7 is not very clear about whether or not it includes sequences from plants ('Chromoviridae').

We agree that the Introduction needed more context on Ty3/gypsy retrotransposons. We have revised it to briefly state that LTR retrotransposons include several major lineages, including Ty1/copia, BEL/Pao, Ty3/gypsy and retrovirus-related elements, and that Ty3/gypsy elements are classified primarily by POL similarity and domain organisation. We also now explain that Ty3/gypsy retrotransposons typically encode GAG and POL proteins, with POL providing the enzymatic activities required for reverse transcription and integration, while noting that ORF arrangement and accessory domains can vary between lineages.

Please note that we stated that our screen did not identify intact env-containing Ty3/gypsy elements in vertebrate genomes that were homologous to the invertebrate errantiviruses analysed here. This is not to say that non-env-containing Ty3/gypsy elements are also absent in vertebrate genomes. Finally, we revised the Figure 7 legend to make clear that the comparison includes representative non-env-containing Ty3/gypsy elements from animals, fungi and plants, including chromovirus or chromovirus-related elements.

(c) The known association of Ty3/gypsi-like retrotransposons from different metazoan phyla with putative membrane fusion proteins (env-like) genes is mentioned in the introduction, but literature information, whether such associations also occur in the context of other retrotransposons (e.g., Ty1/ copia or BEL/pao), is not provided. The abstract is somewhat misleading in this respect. Finally, the different known types of env-like genes are not mentioned and explained as part of the introduction ('env-f', 'envB', 'retroviral env', others?)

We expanded the introduction to introduce literature information of known env-associated retroelements, including Ty1/copia and BEL/pao and explained which ENV types are known to be associated to these elements.

(d) Some key references and reviews might be added:

- Pelisson, A. et al. (1994) https://www.embopress.org/doi/abs/10.1002/j.1460-2075.1994.tb06760.x (next to Song et al. (1994), for the identification of env in Ty3/gypsy)

- Boeke, J.D. et al. (1999) In Virus Taxonomy: ICTV VIIth report. (ed. F.A. Murphy),. Springer-Verlag, New York. (cited by Malik et al. (2000) - for the definition and first use of the term 'errantivirus')

- Eickbush, T.H. and Jamburuthugoda, V.K. (2008) https://doi.org/10.1016/j.virusres.2007.12.010 (on the classification of retrotransposons and their env-like genes)

- Hayward, A. (2017) https://doi.org/10.1016/j.coviro.2017.06.006 (on scenarios of env acquisition)

Thank you. We included these references in the introduction.

(3) Incomplete analysis:

(a) Mobile genetic elements are sometimes difficult to assemble correctly from shortread sequencing data. Did the authors confirm some of their newly identified elements by e.g., PCR analysis or re-identification in long-read sequencing data?

Most newly identified elements are found in contigs/chromosomes that are longer than 100kb. The information of the contig/chromosome size, in which the representative copy of the identified elements are found, can be found in the column “CONTIG_SIZE” in supplementary table S1.

(b) The authors mention somewhat on the side that there are Ty3/gypsy elements with a different arrangement (gag-env-pol instead of gag-pol-env). Why was this important feature apparently not used and correlated in the analysis? How does it map on the RT phylogenetic tree? Which type of env is found with either arrangement? Is there evidence for a loss of env also in the case of gag-env-pol elements?

We agree that the non-canonical GAG-ENV-POL arrangement is an important feature that was insufficiently integrated into the analysis. We have revised the Results and figure annotations to make this clearer. Specifically, we now indicate GAG-ENV-POL elements in the POL tree in Fig S4 and in the HSV/gB-type ENV alignment/architecture figure S8. These elements are found in Nematoda, Bryozoa and Platyhelminthes and all carry HSV/gB-type ENV. They do not form a single monophyletic group in the Pol tree, but instead occur in distinct host-associated clades. They also show different HSV/gBtype cysteine-bridge architectures. Thus, the GAG-ENV-POL arrangement is unlikely to represent a single recent rearrangement event shared by all such elements; rather, it appears to be associated with several deeply diverged HSV/gB-type errantivirus lineages.

We have not inferred specific env-loss events for GAG-ENV-POL elements, because doing so would require a separate analysis of related non-env-containing elements.

(c) Sankey plots are insufficiently explained. How would inconsistencies between trees (recombinations) show up here? Why is there no Sankey plot for the analysis of env-B in Figure 5?

We agree that the Sankey plot was insufficiently explained. We have revised the Figure 4 legend to clarify that the Sankey plot was used as a qualitative visual summary of global congruence between the Pol extended RT/connection phylogeny and the F-type ENV ectodomain phylogeny. We now state that ribbon crossing alone should not be interpreted as recombination, because tree drawings can be rotated without changing topology and the relative order of clades in the two displayed trees may differ. Instead, the relevant signal is whether Pol-defined clades map mostly to corresponding F-type ENV-defined clades. Strong discordance, potentially reflecting recombination, env exchange or poor phylogenetic resolution, would be expected to appear as extensive splitting or many-to-many connections between Pol and ENV clades.

We did not include an equivalent Sankey plot for HSV/gB-type ENV in Figure 5 because we did not construct a global HSV/gB-type ENV phylogeny comparable to the F-type ENV ectodomain tree. Instead, HSV/gB-type ENV proteins were analysed by predicted structural organisation and cysteine-bridge architecture, which are shown in Figure 5 and Supplementary Figure S8.

(d) Why are there no trees generated for env-F and env-B like proteins, including closely related homologous sequences that do NOT come from Ty3/gypsy retrotransposons (e.g., from the eukaryotic hosts, from other types of retrotransposons (Ty1/copia or BEL/pao), from viruses such as Herpesvirus and Baculovirus)? It would be informative whether the sequences from Ty3/gypsy cluster together in this case.

We agree that comparison with homologous fusogens outside Ty3/gypsy retrotransposons is informative. We have therefore added an expanded F-type ENV ectodomain phylogeny that includes representative viral and retroelement-associated F-like proteins, including baculovirus F proteins, paramyxovirus and pneumovirus F proteins, and the BEL/Pao-associated Drosophila Roo F-like protein. In this expanded tree, the added viral sequences formed family-level clades within the broader F-type ENV diversity. Errantivirus F-type ENV proteins did not cluster as a shallow Ty3/gypsyspecific group or as a recent derivative of a single sampled viral family; instead, they spanned a level of diversity comparable to that separating major viral F-protein groups.

For HSV/gB-type ENV proteins, we did not generate an equivalent global phylogeny because the primary sequences and domain organisations of viral class III fusogens and errantivirus HSV/gB-type ENV proteins were too divergent for reliable full ecto domain multiple-sequence alignment. Instead, we added a structural comparison with representative viral and retroelement-associated class III fusogens, including herpesvirus gB, rhabdovirus G, orthomyxovirus GP75/GP64-like proteins, baculovirus GP64 and BEL/Pao-associated gB-like proteins. This analysis showed that viral class III fusogens often retained family-specific cysteine-bridge architectures despite low primary-sequence identity. Errantivirus HSV/gB-type ENV groups showed a comparable pattern, retaining lineage-specific cysteine-bridge architectures despite extensive sequence divergence. We have revised the Results, Methods and supplementary figure legends to clarify these analyses and to distinguish the phylogenetic analysis of F-type ENV from the structural comparison of HSV/gB-type ENV.

(e) Did the authors identify any other env-like ORFs (apart from env-F and env-B) among Ty3/gypsy retrotransposons? Did they identify other, non-env-like ORFs that might help in the analysis? It is not quite clear from the methods if the searches for env-F and envB - containing Ty3/gypsy elements were done separately and consecutively or somehow combined (the authors generally use 'env', and it is not clear which type of protein this refers to).

We agree that this was not sufficiently clear. We have revised the Methods to clarify that the search was designed to identify Ty3/gypsy elements carrying ORFs structurally resembling known envelope/fusogen proteins. In the iterative tBLASTn searches, bait sequences representing both F-type ENV and HSV/gB-type ENV were included together in each round, rather than being searched as two entirely separate pipelines. Candidate elements were then annotated and classified by ORF structure, HHpred/domain similarity and structural prediction.

Among intact Ty3/gypsy candidates recovered by this strategy, we identified two recurrent classes of env-like ORFs: F-type env and HSV/gB-type env. We did not identify an additional recurrent class of env-like ORF among the intact Ty3/gypsy elements analysed here. We also did not identify other recurrent non-env accessory ORFs that were informative for the phylogenetic analyses beyond the GAG, POL and ENV features described in the manuscript.

(f) Why was the gag protein apparently not used to support the analysis? Are there different, unrelated types of gag among non-LTR retrotransposons? Does gag follow or break the pattern of co-evolution between RT and env-F/env-B?

We agree that the role of GAG in the analysis should be clarified. GAG ORFs were used during element annotation to identify intact GAG-POL-ENV or GAG-ENV-POL retrotransposon architectures, but we did not use GAG as a major phylogenetic marker because GAG proteins are less conserved and less reliably alignable across deeply diverged Ty3/gypsy elements than the enzymatic POL domains. Our central question was the acquisition and long-term retention of env-like ORFs by POL-defined Ty3/gypsy retrotransposons. We have revised the Methods to clarify that GAG was used for structural annotation and intactness assessment, whereas phylogenetic analyses were based on the Pol extended RT/connection region and Integrase domain.

(g) Data availability. The link given in the paper does not seem to work (https://github.com/RippeiHayashi/errantiviruses_2025/tree/main). It would be useful for the community to have the sequences of the newly identified Ty3/gypsy retrotransposons listed readily available (not just genome coordinates as in table S1), together with the respective annotations of ORFs and features.

The GitHub repository that contains suggested data is made public. Please check the link again.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Additional Analyses That Could Strengthen Claims (but I concede might be well beyond the scope of this study):

(1) Reconciliation and Gene Tree-Species Tree Analysis: Implementing explicit genetree-species-tree reconciliation (e.g., using Notung or ALE) could formally test the frequency of horizontal transfers vs. vertical inheritance in errantivirus evolution.

We performed targeted host-taxonomy concordance analyses in representative well-sampled clades to address this question, as described in the revised manuscript.

(2) Functional Validation in Non-Insect Hosts: RNA-seq or proteomics from representative non-insect hosts could reveal whether env genes are expressed, and-if possible-experimental assays for envelope function would move beyond computational inference. This is the one thing that might be done in a revision.

We agree that direct functional validation in non-Drosophila species would substantially strengthen the conclusions. Such experiments are beyond the scope of the present revision. However, we now cite our recent work (PMID: 41922845) showing conserved piRNA-mediated defence against errantiviruses across insect orders, providing indirect evidence that these elements remain biologically active outside Drosophila.

(3) LTR Age Dating: Estimating insertion ages using LTR divergence across clades would contextualize the timing of expansion events and help test hypotheses about ancient vs. recent proliferation.

We agree that LTR divergence-based insertion dating would be informative for estimating the timing of recent expansion events. However, our main evolutionary conclusions concern the deeper history of env acquisition and long-term retention across metazoan lineages, rather than the precise insertion age of individual genomic copies. Because the dataset includes elements from highly divergent genomes with variable assembly quality and many multicopy families, systematic LTR dating across all clades would require additional curation and is beyond the scope of the present revision.

(4) Comparative Host Defense Analysis: Surveying host antiviral or transposon defense systems (e.g., piRNA, APOBEC) in lineages rich in errantiviruses could test for signatures of recurrent molecular arms races.

We agree that comparative analysis of host defence pathways, including piRNA and antiviral systems, is an important future direction. However, a systematic survey of host-defence evolution across all errantivirus-rich lineages is beyond the scope of the present revision.

Reviewer #2 (Recommendations for the authors):

Apart from my comments in the Public Review, I have a few additional minor recommendations:

(1) The authors frequently use the term "active" to describe complete retrotransposons. However, in transposon biology, "active" implies recent or ongoing transpositional activity and should therefore be used with caution. Terms such as "complete" or "fulllength" would be more appropriate in this context.

We agree that intact ORF structure should not be over-interpreted as evidence of recent mobilisation. We have therefore revised the manuscript to distinguish element intactness from evidence of recent expansion. Specifically, we now use “intact” or “full-length” to describe element structure. We also clarify that the presence of multiple highly similar copies in the same genome, defined as >98% nucleotide identity across >98% of the three ORFs, is evidence consistent with recent or ongoing genomic expansion, but not definitive proof of current transposition. These changes have been made in the Abstract, Results, Methods and Discussion.

(2) In the third paragraph of the Results, the authors conclude that many identified errantiviruses were mobilized recently. However, since the search strategy specifically targets complete and uninterrupted elements, this may introduce a bias toward younger elements, making the conclusion somewhat circular.

We agree and have revised the wording to avoid over-interpreting completeness as evidence of recent mobilisation. We now distinguish between intact/full-length element structure, which was part of our search strategy, and independent evidence for recent or ongoing mobilisation, such as the presence of multiple highly similar copies. We have therefore softened statements that previously implied that intact ORFs alone demonstrate recent activity.

(3) The first paragraph of the Introduction requires appropriate references.

We have now added references to enhance the readability of the first paragraph of the introduction.

(4) It would benefit readers if the retrotranspositional process were briefly explained, including a description of what the tRNA primer binding site (PBS) is and its role in retrotransposition.

In the same part of the introduction, we now describe the role of the tRNA primer binding site in retrotransposition.

Reviewer #3 (Recommendations for the authors):

Suggestions and requests for clarification currently are part of the public review, as they also point the reader to critical open questions if the authors decide not to amend their version of the manuscript.

Version 2

Abstract

Retrotransposons or endogenous retroviruses (ERVs) essentially carry open reading frames of gag and pol, which are utilized to selfishly replicate themselves in the host germline genome. One rare example of ERVs that additionally carry envelope genes is Ty3/gypsy errantiviruses in Drosophila. Though they are structurally analogous to retroviruses, it remained unclear whether envelope-containing Ty3/gypsy elements represent recent, lineage-specific acquisitions of viral fusogens or an ancient association between retrotransposons and envelope-like genes. We systematically searched for intact envelope-containing ERVs that are homologous to Ty3/gypsy in invertebrate metazoan genomes and found that they are widespread across taxa including ancient animals such as cnidarians, ctenophores and tunicates. Many elements occur as multiple highly similar copies in their respective genomes, consistent with recent genomic expansion in some host lineages. Envelope genes are classified into those that resemble glycoprotein F from paramyxoviruses and glycoprotein B from herpesviruses, and both types are equally abundant and widespread. Phylogenetic and structural analyses revealed that envelope genes have largely diverged with pol genes as well as with the host organisms throughout their evolutionary history and recombined infrequently, suggesting that the envelope acquisition to ERVs is ancient and likely dates to before the split of bilaterian and non-bilaterian animals in Pre-Cambrian era.

Introduction

Long terminal repeat (LTR) retrotransposons are one of the most abundant classes of transposons that are ubiquitously present in eukaryotic genomes (Havecker et al. 2004; Wicker et al. 2007). They replicate by reverse-transcribing the messenger RNA and inserting the DNA copy into the host genome (Wilhelm and Wilhelm 2001). They essentially carry two open reading frames (ORFs), namely, gag and pol. gag encodes for a structural protein that packages RNA into capsids while pol encodes a multi-functional protein that contains the reverse transcriptase, RNase H and integrase domains. Gag and pol ORFs are flanked by 5’ and 3’ LTRs that contain regulatory sequences required for transcription and replication. Reverse transcription is primed from a tRNA primer-binding site located near the 5’ LTR. Although intact copies can continue to replicate, individual insertions acquire mutations and degrade over time once they become inactive. Some retrotransposon remnants remain in the genome to provide functional ORFs and regulatory sequences of host gene expression (Chuong et al. 2017).

LTR retrotransposons are divided into several major lineages, including Ty1/copia, BEL/pao and Ty3/gypsy elements, which are distinguished primarily by POL sequence similarity and domain organisation (Eickbush and Jamburuthugoda 2008). Ty3/gypsy retrotransposons are especially closely related to retroviruses among LTR retroelements and typically encode GAG and POL proteins, with POL providing the enzymatic activities required for reverse transcription and integration. They are broadly distributed across eukaryotes, including animals, fungi and plants, although their abundance, domain architecture and ORF arrangement vary substantially between lineages. In this study, we use “Ty3/gypsy retrotransposon” for elements whose POL proteins classify with the Ty3/gypsy group, as described in the Methods.

Retroviruses are closely related to LTR retrotransposons, but differ from most retrotransposons by carrying an envelope gene, env, which encodes a membrane fusion protein essential for infectivity (Xiong and Eickbush 1988; Xiong and Eickbush 1990). Although they typically horizontally transmit between individual animals, they can occasionally infect the germline cells of the host, vertically transmit thereafter, and become known as endogenous retroviruses (ERVs). A well-studied example of such endogenization events is found in Koala, in which nearly identical copies of an exogenous lymphoma-causing retrovirus are found in the genomes of geographically distinct Koala populations (Tarlinton et al. 2006; Zheng et al. 2020). Although individual ERV insertions are subject to genetic drift and degrade over time after they acquire mutations, some insertions remain infectious and continue to make new insertions in the germline genome (Ribet et al. 2008; Miyazawa et al. 2010; Anai et al. 2012; Verneret et al. 2025). However, it is unclear how long such infectious ERVs vertically transmit in a long evolutionary time. When homologous ERVs are found in distantly related species, it can be either seen that they have been inherited from their common ancestor or recently transmitted from one species to another through horizontal transfer. The mechanism of such horizontal transfer events is even less clear although it is often hypothesised that exogenous retroviruses that are borne by one species infect the other species that shares the habitat (Hayward et al. 2020; Simpson et al. 2022; Mottaghinia et al. 2024).

Envelope-like genes are not restricted to vertebrate retroviruses. Canonical vertebrate retroviruses generally encode retroviral class I Env proteins, whereas some ancient retrovirus-like lineages, such as lokiretroviruses, encode Env proteins related to Mononegavirales-like fusion glycoproteins, including paramyxovirus/pneumovirus-like F proteins (Wang and Han 2022). env-like ORFs have been reported in several retroelement contexts, including Ty3/gypsy, Ty1/copia-like elements, and BEL/Pao elements such as the Drosophila element Roo, and mosquito BEL/Pao elements associated with Chuviridae-like glycoproteins (Malik et al. 2000; de la Chaux and Wagner 2009; Dezordi et al. 2020). Their encoded proteins can resemble unrelated viral fusogens. Known env-like proteins associated with invertebrate retroelements include glycoprotein F-like proteins, here referred to as F-type ENV, and glycoprotein B-like class III fusogens, here referred to as HSV/gB-type ENV. These proteins are structurally analogous to viral entry proteins, but their evolutionary origins and the extent to which they mediate infectivity in retroelements remain unclear (Hayward 2017).

The best-characterised env-containing retroelements in invertebrate genomes are Drosophila gypsy elements (Pélisson et al. 1994; Song et al. 1994; Terzian et al. 2001; Pelisson et al. 2002). These so-called errantiviruses behave like bona fide retroviruses, making virus-like particles in the ovarian somatic cells and infecting the germline cells (Song et al. 1997; Leblanc et al. 2000). Therefore, errantiviruses replicate within the genome of the same animal as other retrotransposons do but through an intra-gonadal infection mechanism. Errantiviruses are widespread in Drosophila genomes and form a monophyletic group among other Drosophila gypsy elements based on the phylogeny of the pol ORF, suggesting that a subtype of gypsy acquired the envelope gene in the common ancestor of Drosophila and has been vertically transmitting thereafter (Bargues and Lerat 2017). It is postulated that the ancient gypsy retrotransposon in Drosophila acquired the envelope gene from an insect-specific DNA virus, baculovirus, through recombination (Malik et al. 2000; Rohrmann and Karplus 2001). However, Ty3/gypsy LTR retrotransposons with envelope genes have also been sporadically documented in non-insect animals, such as in nematodes, rotifers, and tunicates, obscuring their true origin (Volff et al. 2004; Rodriguez et al. 2017; Sacco et al. 2022).

Here we comprehensively searched for envelope-carrying Ty3/gypsy retrotransposons in invertebrate metazoan genomes. We found elements with uninterrupted ORFs flanked by putative LTR sequences in nearly every major phylum, including ancient animals such as cnidarians, ctenophores, and tunicates. Furthermore, many of these elements occur as multiple highly similar copies in their respective genomes, consistent with recent or ongoing genomic expansion in some host lineages. Phylogenetic and structural analyses of the pol and envelope genes show that they have co-evolved with their host species for a long evolutionary time. Therefore, we propose that the acquisition of the envelope gene to retrotransposons occurred anciently in metazoans, and they have been vertically transmitting since pre-Cambrian era.

Results

The approach to study the phylogeny and evolution of LTR retrotransposons, is conventionally to mine the genome for sequences which share homology to retroviral gag, pol, or envelope (env) genes. Prior studies on ERVs and retrotransposons have assembled trees based on reverse transcriptase (RT) or Integrase (IN) from pol to obtain insights (Xiong and Eickbush 1990; Wang and Han 2021). One can also investigate synteny of loci between different hosts, estimating an age of origin for those elements (Bonnaud et al. 2005; Ueda et al. 2020).

This study aims to investigate the evolution of env genes acquired by intact Ty3/gypsy retrotransposons, and to explore structural variations within those elements. The conventional approach, therefore, needs to be altered to approach this. We only consider intact copies that retain uninterrupted open reading frames (ORFs) for GAG, POL and ENV, because non-degraded ORFs provide more reliable material for domain annotation, structural prediction and phylogenetic comparison (Fig 1A). Because intact ORF structure was part of our search strategy, the recovery of intact ORFs should not by itself be interpreted as evidence of recent mobilisation. Where possible, we also recorded whether closely related copies were present in the same genome, because multicopy structure provides additional evidence consistent with recent or ongoing genomic expansion. The study of syntenic loci here is inappropriate, as the genomic location of recently mobilised transposons will likely never be the same in evolutionarily distant hosts. Like most studies of repeat elements, the sequences are highly diverse, and we therefore supplement phylogenetic evidence with structural predictions.

The widespread nature of intact GAG-POL-ENV elements across metazoan genomes.

(A) Open reading frame arrangement of intact errantivirus copies. GAG encodes the structural capsid-like protein, POL encodes the enzymatic protein containing protease (PRO), reverse transcriptase (RT), RNase H and Integrase (IN) domains, and ENV encodes the envelope-like fusogen. tRNA primer binding sequence is used for reverse transcription. (B) Summary of Errantiviruses found in Metazoan phyla with type of ENV protein indicated. (C) Copy number of intact errantivirus copies in metazoans. Column represents a genome assembly, each tile representative of annotated errantivirus, colour of tile represents copy number.

We systematically searched for homologous elements of Dipteran Ty3/gypsy errantiviruses in the referenced genomes across metazoan phyla using the annotated protein coding sequences of gag, pol and env for tBLASTn searches (see methods and Fig. S1). The first tBLASTn search identified elements outside Diptera, therefore, we extended the search using the representative of newly identified elements as baits. We repeated this process and examined whether the additional baits uncovered elements that were not represented in the previous tBLASTn searches by identifying new clades in the phylogenetic tree of POL (see methods and Fig. S1). After five iteration of tBLASTn searches, we no longer discovered elements that formed a new POL phylogenetic clade (Fig. S1B). These steps allowed us to obtain a comprehensive set of elements in metazoan genomes that are homologous to Dipteran Ty3/gypsy errantiviruses (Table S1, summarised in Fig 1). The majority of the newly identified genomic copies carry putative LTRs and tRNA primer binding sequences (1151 out of 1512 elements, Table S1), suggesting their recent integrations. Additionally, most of the annotated elements sit within contigs larger than 0.1 mega bases (1434 elements, Table S1), indicating their true origins in their respective genomes. We generated maximum-likelihood phylogenetic trees from multiple sequence alignments of two conserved Pol regions: the Integrase (IN) domain and an extended RT/connection region. The extended RT/connection region includes the RT polymerase core and the downstream connection subdomain, and is therefore broader than the RT polymerase core alone but does not include the complete canonical RNase H domain analysed structurally in Figure 6. ORFs of annotated Drosophila gypsy errantiviruses were also included in the trees to aid classifications.

We performed a similar analysis on vertebrate genomes using known vertebrate retroviral gag, pol, and env sequences as bait in addition to Drosophila errantiviruses. We did not identify any intact ERVs in vertebrate genomes that were homologous to invertebrate elements, nor did we find elements in invertebrate genomes that predicted homology to vertebrate elements. There is therefore a clear phylogenetic separation between invertebrate elements and vertebrate ERVs, despite structural similarities. The investigation on the vertebrate ERVs will be reported elsewhere, as it is outside the scope of the present study.

For clarity, we use the following terminology in this study. ERV refers to an endogenous retrovirus or retrovirus-like LTR retrotransposon present in the host genome. Ty3/gypsy retrotransposon refers to an LTR retrotransposon classified with the Ty3/gypsy group based on POL similarity (see methods). Errantivirus is used provisionally and operationally to refer to env-containing Ty3/gypsy retrotransposons identified in this study, following the historical use of this term for env-containing Metaviridae/Ty3/gypsy elements (Boeke et al. 2000). This usage is not intended as a formal taxonomic proposal and does not imply that all such elements share the experimentally characterised Drosophila gypsy lifecycle. We use lower-case env for the gene and upper-case ENV for the encoded protein. F-type env/ENV refers to glycoprotein F-like env genes or proteins, whereas HSV/gB-type env/ENV refers to herpesvirus glycoprotein B-like env genes or proteins.

Multiple intact elements of env-carrying Ty3/gypsy retrotransposons are found widespread across metazoan species

We found that intact copies of errantiviruses carrying all three retroviral ORFs are found not only outside of insect orders but are ubiquitously present across most major metazoan phyla. This includes species from ancient phyla such as Cnidaria, Tunicates and Ctenophora, but excludes Echinodermata and Porifera (Fig 1B and Table S1). To test whether the apparent absence of env-carrying Ty3/gypsy elements in Porifera and Echinodermata could be explained by poor assembly quality or a general failure to recover full-length retrotransposons, we searched two representative genomes from each phylum for full-length GAG-POL Ty3/gypsy elements using the same general annotation strategy. We identified multiple full-length, multicopy GAG-POL elements in all four genomes examined, many of which were flanked by predicted LTR sequences and associated with putative tRNA primer-binding sites (Table S3). Thus, the lack of detectable GAG-POL-ENV elements in these representative Porifera and Echinodermata genomes is unlikely to be explained simply by an inability to recover intact Ty3/gypsy-like retrotransposons from these assemblies. Many carry ENV proteins that are homologous to glycoprotein F, commonly found in paramyxoviruses, such as Nipah, Hendra and Respiratory Syncytial Viruses (McLellan et al. 2013; Dang et al. 2019), but is also carried in insect baculoviruses (and Drosophila errantiviruses) (Malik et al. 2000; Rohrmann and Karplus 2001). Other elements we identified instead carry ENV that resembles glycoprotein B (gB), famously found as part of the glycoprotein complex in Herpes Simplex Viruses (HSV). Both types of ENV glycoproteins are carried by errantiviruses found in ancient hosts such as rotifers, tunicates, cnidarians and ctenophores, as well as in hosts from insect and arthropod orders (Fig 1B).

Though, errantiviruses found in Nematoda, Bryozoa and Platyhelminthes only carry HSV/gB-type ENV, and the ORFs are arranged as GAG-ENV-POL rather than GAG-POL-ENV. These GAG-ENV-POL elements do not form a single monophyletic group in the POL phylogenetic tree, but instead occur in distinct host-associated clades, as shown later in Supplementary Figure S4B and Supplementary Figure S8. Most elements that were identified have multiple intact copies in their respective genomes (as defined as copies with >98% nucleotide identity across >98% of the three ORFs, see methods, Table S1), a hallmark feature of recent transposition (Cordaux and Batzer 2009; Gagnier et al. 2019), indicating that most elements that we have annotated likely behave as bona-fide retrotransposons (Fig 1C). Activity of these elements are highly varied, as the copy number can be as little as one intact copy, to hundreds of intact copies in the genome (Fig 1C, Table S1). Importantly, multi-copy ERVs are present in nearly all animal phyla where we see intact elements, suggesting that these elements have been recently transposing in a wide range of metazoan species.

The phylogenetic tree assembled on the intact errantivirus pol genes shows congruence, meaning that the tree assembled on extended RT/connection region closely reflects that of IN. The simplified tree in Figure 2 is used to summarise the major phylogenetic groups; the complete POL extended RT/connection tree, the corresponding IN tree and their congruence are shown in Supplementary Figure S2. Errantiviruses within most clades in one tree are represented in the same or closely related clades in the other (Fig. S2B). Importantly, the errantiviruses found in phyla outside of insects, are in a different clade to those found within, which we call “ancient errantiviruses” and “insect errantiviruses”, respectively (Fig 2 and S2A). Notably however, few elements found in insects sit within clades of ancient errantiviruses, rather than other insect ones (Fig 2). Annotated Drosophila gypsy elements included in the tree, namely, Gypsy-ZAM, Gypsy-Gypsy/Gypsy-5_DGri and Gypsy-DM176/Idefix present themselves in different clades while they cluster together with other insect errantiviruses, in an arrangement with previous studies that have classified Drosophila gypsy elements (Bargues and Lerat 2017) (Fig 2). Though, one element from a cnidarian species Paleozoanthus reticulatus clusters together with insect elements, in a clade adjacent to DM176 and Idefix (Fig 2), its origin is unclear as the element is found on a short genomic contig (Table S1).

Phylogeny of POL extended RT/connection region from intact metazoan errantiviruses.

Maximum likelihood tree constructed on representative POL extended RT/connection regions of intact errantiviruses (see methods) and rooted to the midpoint. The “ancient errantiviruses” and “insect errantiviruses” phylogenetic groups are highlighted in green and white respectively. Positions of known dipteran errantiviruses, Gypsy_ZAM (1), Gypsy_Gypsy5_Dgri (2), Gypsy_Gypsy (3), Gypsy_Idefix (4), and Gypsy_DM176 (5) are indicated. Host classifications of errantiviruses are annotated in colour, non-arthropod Protostomia (yellow), Deuterostomia (red), Diploblasta (orange), non-insect Arthropoda (green), Insecta (blue). One cnidarian element within the group of insect errantiviruses is marked by an asterisk.

Many of the newly identified “insect errantiviruses” sit in separate clades to the Annotated Drosophila gypsy elements (Fig 2, tree positions I3 – I6 in Fig. S2A). These “insect errantivirus” clades are strongly supported by bootstrap values greater than 95 and are additionally supported by the annotations of the tRNA primer binding site (PBS) found in each element. For example, Tree position I1 within “insect errantiviruses” mainly carry tRNASer, whereas Tree position I2, consisting of only dipteran elements, predominantly use tRNALys, though few use tRNASer or tRNAPro (Fig. S2A and S2C). Elements in insects within positions I4, I5 and I6 predominantly carry tRNALeu primer binding sites, in addition to Serine and other uncommon tRNAs (Fig. S2C). This is mostly consistent with past works, that have identified Drosophila gypsy elements to carry either Serine or Lysine tRNA PBS (Terzian et al. 2001; Stefanov et al. 2012). Notably, most discrete clades comprised elements from the same insect orders, suggesting long term vertical transmission and divergence within related species (Fig. S3). A detailed view of the insect errantivirus POL extended RT/connection clades, including host order, ENV type and the positions of phylogenetic groups I1–I6, is provided in Supplementary Figure S3.

The “ancient errantivirus” group also consists of clades strongly supported by bootstrap values greater than 95, though the tRNA PBS content within the clades is much more divergent (Fig. S4C). Notably however, errantiviruses found in insects that sit within this group, in tree positions A1, A2 and A7 continue to carry predominantly tRNASer, tRNALys and tRNALeu (Fig. S4B). Discrete clades are consistently made up of elements in hosts from the same, or closely related phyla (Fig. S4B), for example, tree position A5 consists of elements found in Thecostraca, Annelida and Mollusca, each of which form their own discrete clade (Fig. S4B). This indicates that they were either present in their common ancestor or inherited more recently within the same animal group. Given that neighbouring clades consist of elements found in related animal phyla, it is more likely that they are ancient in nature. For example, clades that consist of insect elements are flanked by clades of other arthropod elements (Fig. S4B, tree position A2). Clades containing Annelida elements often cluster with or next to mollusc elements (Fig. S4B, tree positions A5-A7). This congruence between clades and species is highly indicative of long-term vertical transfer, as known in non env-containing retrotransposons (Sormacheva et al. 2012; Wanner and Faulk 2021). Within larger clades containing a wider variety of host orders such as tree position A9, the separation is still evident and strongly supported clades only consist of elements from the same order of host (Fig. S4B). Supplementary Figure S4 provides the full phylogenetic context for the “ancient errantivirus” group, including host taxonomy, ENV type, tree positions A1–A11 and tRNA PBS annotations.

To test the host-taxonomic structure more directly, we performed targeted permutation analyses in two well-sampled POL extended RT/connection subclades in tree positions I1 (Fig. S3) and A6 (Fig. S4B), consisting of elements from Lepidoptera and Annelida, respectively (Fig. S5). We scored internal nodes with bootstrap support of ≥95 and 3–10 descendant tips, then randomly permuted host-taxonomic labels across the fixed tree topology while preserving the observed label counts. In the Annelida clade, supported small subclades were far more host-concordant than expected by chance: all six scored subclades consisted of elements from the same host family, compared with a null expectation of 0.27 subclades, and four of six consisted of elements from the same host species, compared with a null expectation of 0.13 subclades (10,000 permutations; empirical P = 1.0 × 10-4 for both metrics). The Lepidoptera clade showed a more mixed pattern, but several highly supported subclades were enriched for related host groups at the superfamily or broader taxonomic level. These targeted analyses do not exclude rare horizontal transfer, particularly between closely related hosts, but indicate that host labels are not randomly distributed across local POL phylogenetic topologies; the analysed Annelida A6 and Lepidoptera I1 subclades are shown in Supplementary Figure S5.

Furthermore, clades formed on the trees using pol proteins tend to carry the same type of env (Fig 3A and S4A). For example, within A6 in the “ancient errantivirus” group, elements that carry HSV/gB-type env are in a separate clade to those carrying F-type env (Fig. S4B). Though some clades appear to be a mix of both, elements with HSV/gB-type env still tend to cluster together, indicating that they have stayed with pol for long evolutionary periods of time.

env classifications of intact metazoan errantiviruses.

(A) Shown is a POL extended RT/connection maximum likelihood tree of the representative errantiviruses (see methods) with host (*) and type of ENV protein (**) indicated. Colours represent non-insect animals (*) (yellow), insects (*) (blue), glycoprotein F (F) (**) (orange), and glycoprotein B (HSV/gB) (**) (purple). Clades containing hymenopteran elements in “Insect errantiviruses” that carry either F or HSV/gB env are annotated as 1 and 2. (B) Close up of clades 1 and 2 in “Insect errantiviruses”. F-type and HSV/gB-type errantiviruses that are both found in Lasioglossum lativentre are marked by asterisks.

ENV glycoproteins carried by errantiviruses are highly structured, and their divergence follows that of pol

As previously mentioned, we found that errantiviruses can carry HSV/gB-type or glycoprotein F-type env. While most HSV/gB-type env is found in the “ancient errantivirus” clade, one clade in “insect errantiviruses” exclusively consists of hymenopteran errantiviruses which carry both HSV/gB-type and F-type env genes (Fig 3A-B). Interestingly, elements with HSV/gB-type env and F-type env separate at the base of this clade while they both are found in a wide range of species superfamily within Hymenoptera, suggesting that both HSV/gB-type and F-type env have been together with their pol genes at least throughout the evolution of hymenopteran species (Fig 3B).

To further characterise the ENV proteins carried by errantiviruses, we identified the ectodomains within the env ORF, which is flanked by signal peptides and a transmembrane domain, using DeepTMHMM (https://doi.org/10.1101/2022.04.08.487609). We then analysed trimeric configurations of these ectodomains from representative errantiviruses from the phylogenetic tree using Alphafold 2 (Jumper et al. 2021).

F glycoproteins analysed are highly structured and consistently form four domains that are highly conserved across all elements (Fig 4A). They also predicted a highly conserved Furin cleavage site and hydrophobic fusion peptides in the ectodomain (Fig 4A and S6). The alignment-level evidence supporting conservation of the predicted furin cleavage site and hydrophobic fusion peptide is shown in Supplementary Figure S6. The 3D fold of the ectodomain structures forms a “head” (domains 1-3) and a “stalk” (domain 4), strongly resembling 3D structures of F ectodomains characterised in previous studies, such as those from RSVs (McLellan et al. 2013) (Fig 4C). Most errantiviruses also have an α helix at the C terminus of the ectodomain, referred to as “Domain A” (Fig 4A and 4C). The predicted 3D trimeric structure also reveals two potential configurations of the glycoprotein, based on domain 4: either a long α helix, or two shorter helices. These long and short barrel structures resemble pre– and post-fusion structures of glycoprotein F described in prior studies (McLellan et al. 2013). These highly structured domains in the intact env ORFs in these errantiviruses demonstrates that env has remained intact and could suggest that their function also remains intact as well.

Structure and phylogeny of glycoprotein F-type ENV found in intact errantiviruses.

(A) Domain arrangement of glycoprotein F ectodomain in Errantiviruses. (B) A Sankey plot of POL extended RT/connection and glycoprotein F-type ENV ectodomain maximum likelihood trees for representative errantiviruses. In the Sankey plot, each ribbon links the clade assignment of the same element in the Pol extended RT/connection tree and the F-type ENV ectodomain tree. The plot is intended as a qualitative summary of global congruence between the two trees. Ribbon crossing can result from the relative ordering or rotation of clades in the two tree drawings and should not by itself be interpreted as recombination. Instead, broad one-to-one mapping between Pol-defined and ENV-defined clades indicates concordance, whereas extensive splitting or many-to-many connections indicates discordance consistent with recombination, env exchange or differential phylogenetic resolution. Bootstrap values for nodes are given. The “ancient errantivirus” group is highlighted in green. (C) Monomeric and trimeric structures of the “pre-fusion configuration” and the “post fusion configuration”. Schematic cartoons of F-type ENV ectodomain domain 4 are shown in (C’) and (C’’) Respectively. Structures are taken from Diptera_Sciomyzidae_Cmar_errantivirus_19 and Branchipoda_Diplostraca_Dcar_errantivirus_3 for “pre fusion” and “post fusion” structures in (C).

To explore the phylogenetic relationships between pol and env-F, we generated multiple sequence alignments of the ectodomains and POL extended RT/connection regions from representative elements and assembled phylogenetic trees. We found that the ENV ectodomains cluster similarly to the POL extended RT/connection regions, and ENV proteins tend to form clades within the same or closely related species as they do in the POL tree (Fig 4B). Importantly, the ENV ectodomains found in “ancient errantiviruses” cluster separately to those found in “insect errantiviruses”, including those insect elements that present themselves with ancient errantiviruses in the POL tree (Fig 4B). This suggests that env-F genes have stayed with pol, and the acquisition of env likely occurred anciently.

To further compare errantivirus F-type ENV proteins with related fusogens outside Ty3/gypsy retrotransposons, we repeated the F-type ENV ectodomain phylogeny after adding representative viral and retroelement F-like proteins, including baculovirus F proteins, paramyxovirus and pneumovirus F proteins, and the BEL/Pao-associated Drosophila Roo F-like protein (Fig. S7). The added viral sequences formed family-level clades within the broader diversity of F-type ENV proteins. The inclusion of these external homologues reduced the apparent separation between insect and ancient errantivirus F-type ENV proteins, indicating limited resolution for some deep relationships in the F-type ENV tree. Nevertheless, the errantivirus F-type ENV proteins did not collapse into a shallow Ty3/gypsy-specific group, nor did they cluster as a recent derivative of a single sampled viral family. Instead, their diversity was comparable in scale to that separating major viral F-protein groups, consistent with deep diversification of F-type ENV proteins associated with errantiviruses.

The ectodomain of errantiviruses that carry HSV/gB-type env are either flanked by signal peptides and a transmembrane domain, or by two transmembrane domains and the Furin cleavage site is not always conserved (Fig 5A). Like F, the trimeric structures of HSV/gB-type ENV in errantiviruses are also highly structured. They carry five highly conserved and distinct domains (designated as domain I – V) arranged in a hairpin configuration, that fold into a trimeric structure with a base, stalk and crown (Fig 5B). Additionally, domain I carries highly conserved “fusion loops” consisting of hydrophobic amino acid residues. These domains have also been characterized in Herpes Simplex Viruses in prior studies (Heldwein et al. 2006; Connolly et al. 2011). Notably, errantiviruses found in insect orders that carry HSV/gB-type env, only carry one of these fusion loops (Fig. S8). The domain I–II alignments underlying these cysteine-bridge classifications are shown by host group in Supplementary Figure S8, allowing conservation and variation of cysteine positions to be compared across HSV/gB-type ENV lineages. The structure of the HSV/gB-type ENV carried by these errantiviruses are stabilized by disulphide bonds between cysteine residues, which we refer to as cysteine bridges (Lopper and Compton 2002; Vollmer et al. 2020). These bridges are highly conserved but appear to vary the most within domains I and II of the HSV/gB ectodomain (Fig 5C and S8). Additionally, in some cases, a cysteine bridge within domain I is located near the fusion loops (Fig 5C and S8).

Structure of glycoprotein HSV/gB-type ENV found in intact errantiviruses.

(A) Structural arrangements of Ectodomains, Transmembrane domains and signal peptides found in HSV/gB-type errantiviruses. (B) Shown are the arrangement of domains of the HSV/gB-type ENV ectodomain, and Alphafold 2-predicted 3D monomer and trimer structures of the ectodomain. The structure is taken from Hemiptera_Pemphigidae_Elan_errantivirus_1. (C) Fusion loops and conserved cysteine bridges on structure of the errantivirus HSV/gB-type ENV ectodomain. Shown are variable cysteine bridges in the errantivirus HSV/gB-type ENV ectodomains, in domain I and II (Indicated with orange bar. Proximity of cysteine bridge to fusion loops of domain I are indicated for structures within C1, C2 and C3. Structures taken from Cnidaria_Hydrozoa_Chem_errantivirus_5, Bryozoa_Gymnolaemata_Cpal_errantivirus_2, and Tunicata_Ascidiacea_Clep_errantivirus_1, for C1, C2 and C3 configurations, respectively. (D) Shown is the summary of the cysteine bridge configurations found in HSV/gB-type ENV of different metazoan orders. Positions for cysteine bridge configurations within Nematoda and Platyhelminthes shown on Simple schematic. Exact positions of all cysteine bridges can be found in Supplementary Figure S8.

RNase H domains found in the bridge region of intact errantiviruses.

(A) Alphafold 2-predicted structures of domains found in the bridge regions of errantiviruses, with topology maps of α helices (grey) and β sheets (red). Structures are taken from Annelida_Polychaeta_Apac_errantivirus_7, Bryozoa_Gymnolaemata_Cpal_errantivirus_3, Coleoptera_Coccinellidae_Npum_errantivirus_1, Annelida_Polychaeta_Slim_errantivirus_9, Diptera_Cecidomyiidae_Smos_errantivirus_1 and Hymenoptera_Apoidea_Obic_errantivirus_2 for RNase H/Tether, ‘mini’ RNase H, ‘pseudo’ RNase H domains, “Mini” domain 2, “Pseudo” domain 2 and “Pseudo” domain 3, respectively. (B) Shown is the POL extended RT/connection maximum likelihood tree of representative errantiviruses, with RNase H domain architecture in bridge region of pol highlighted. “Insect errantiviruses” with RNase H + pseudo–RNase H (blue), “insect errantiviruses” with bridge region outliers (dark blue), “ancient errantiviruses” with RNase H + mini domain (green), “ancient errantiviruses” with bridge region outliers (yellow). (C) The structure of the bridge region of yeast Schizosaccharomyces pombe Ty3 element tf1 predicted by Alphafold 2.

To determine whether conservation of cysteine-bridge architecture simply reflected high primary-sequence similarity, we compared errantivirus HSV/gB-type ENV proteins with representative viral and retroelement-associated class III fusogens, including herpesvirus gB, rhabdovirus G, orthomyxovirus GP75/GP64-like proteins, baculovirus GP64 and BEL/Pao-associated gB-like proteins (Backovic and Jardetzky 2009) (Fig. S8). We did not infer a global phylogeny for these proteins because their primary sequences and domain organisations were too divergent for reliable full-ectodomain alignment. Instead, we compared predicted domain organisation, cysteine-bridge architecture and pairwise amino-acid identity within each viral family or errantivirus ENV group.

This comparison showed that conserved cysteine-bridge architecture can persist despite extensive primary-sequence divergence. For example, the sampled herpesvirus gB proteins, representing alpha-, beta– and gammaherpesvirus homologues, shared only 27.8–30.3% pairwise amino-acid identity across aligned residue–residue positions while sharing an identical cysteine-bridge architecture (Burke and Heldwein 2015) (Fig. S8 and Table S4). Rhabdovirus G proteins and orthomyxovirus GP75/GP64-like proteins showed similarly low pairwise identities, 21.8–27.7% and 24.3–30.1%, respectively while also retaining mostly shared cysteine-bridges (Belot et al. 2020). For clarity, we use labels such as C1a and C2b to denote recurrent HSV/gB-type cysteine-bridge configurations defined in Figure 5 and Supplementary Figure S8. These labels describe ENV structural architecture rather than formal phylogenetic or taxonomic groups. Several errantivirus HSV/gB-type ENV groups showed comparable or lower levels of primary-sequence identity while retaining lineage-specific cysteine-bridge architectures, including Cnidaria elements with the C1c configuration (18.1–32.2%), Ctenophora elements with the C1c configuration (21.5–29.5%), Bryozoa elements with the C2b configuration (median 28.1%), Branchiopoda elements with the C2c configuration (median 27.1%), Tardigrada elements with the C2c configuration (median 26.5%) and Nematoda elements with the C1a/d configuration (median 19.9%). Thus, conserved cysteine-bridge architecture in errantivirus HSV/gB-type ENV groups cannot be explained simply by recent duplication or high overall sequence identity. Rather, it supports long-term structural constraint on lineage-specific ENV architectures despite extensive primary-sequence divergence. These pairwise identities are used here as descriptive measures of sequence divergence, not as calibrated molecular-clock estimates.

The configuration of cysteine bridges appears to be specific to phyla, though this does not seem to be reflected by their positions in the tree based on POL extended RT/connection tree. For example, Bryozoa elements (Tree positions A3 and A9 in fig S4B) all carry configuration ‘C2b’, despite being found in different clades on the tree, even though they each cluster themselves separately from other elements (Fig. S4B, 5D and S8). Similarly, errantiviruses carrying HSV/gB-type env found in insect phyla mostly have the same configuration regardless of their position in the POL extended RT/connection tree. For example, insect elements within the “ancient errantivirus” group, and insect elements within the “insect errantivirus” group, all have the same cysteine bridge configuration in HSV/gB (Fig 5D and S8). There are, however, cases where more than one type of configuration can be found within the same phyla, namely Nematoda and Platyhelminthes (Fig 5D). Though, in those cases, they appear to continue to separate based on POL extended RT/connection tree (Fig. S8). For example, Nematoda elements with configuration ‘C2a’ and ‘C2c’ are found in a different clade to those with ‘C1a’ and ‘C1d’ (Fig 5D). Although, it should be noted that elements with ‘C1a’ in Nematoda are found in two different clades (Fig 5D).

As mentioned previously, the clades that form on POL extended RT/connection mostly separate elements that carry HSV/gB-type ENV from F-type ENV within discrete clades (Fig. S4B). This, with the fact that the same substructures can be found in different orders, for example in Nematoda, Bryozoa and Platyhelminthes, suggests that their origin is likely ancient. The overall specificity of certain residues to host orders further suggests that recombination events would have only occurred between elements within the same animal order. Furthermore, the overall separation of distinct cysteine bridge configurations of HSV/gB-type ENV, that are represented in the POL extended RT/connection tree within ancient phyla, suggests that those recombination events occurred anciently. While insect elements are separated on the tree, ERVs with HSV/gB-type ENV tend to cluster separately from those with F-type ENV. For example, in the “insect errantivirus” group, the clade that almost exclusively carries HSV/gB-type ENV sits sister to another clade that only carries F-type ENV (Fig 3B). Not only do they separate from each other, but HSV/gB-type ENV can be found in a variety of superfamilies within Hymenoptera, which suggests their ancient origin (Fig 3B). Some species, such as a hymenopteran Lasioglossum lativentre, carry elements with HSV/gB-type, as well as those with F-type ENV (Fig 3B and Table S1). Despite being found in the genome of the same species, the pol of errantiviruses with F– and HSV/gB-type ENV do not stay together, further supporting their more ancient origin (Fig 3B). It should be noted that some clades, such as ones in tree position A2 within “ancient errantiviruses”, have elements that carry F– and HSV/gB-type ENV sitting together within the same discrete clade, suggestive of more recent recombination (Fig. S4B).

The RNase H-containing bridge region of pol separates “insect errantiviruses” from “ancient errantiviruses”

To further characterise the errantiviruses identified in this study, we next asked whether Pol domain architecture differed between the major phylogenetic groups. We focused on the RNase H-containing bridge region, which we define here as the region spanning the canonical RNase H domain and the C-terminal region between RNase H and Integrase. This region is distinct from the extended RT/connection region used for the Pol phylogenetic analyses above. The connection subdomain included in the extended RT/connection alignment corresponds to a partial RNase H-like fold, whereas the canonical RNase H domain analysed here contains the complete β1–β2–β3–α–β4–α–β5–α architecture shown in Figure 6A. Because RNase H-related domains have changed during retroelement evolution, including the duplicated and partially degraded RNase H-like tether/connection domain described in vertebrate retroviruses (Malik and Eickbush 2001), the bridge region provides an additional structural feature for comparing errantivirus lineages. We therefore characterised the bridge region of representative elements using AlphaFold 2, focusing on the organisation of the canonical RNase H domain and the downstream RNase H-like or mini-domain structures.

We found various structures within the bridge region, mostly consisting of core β-sheets resembling the 3D fold arrangement of RNase H domains but we also found other uncommon substructures (Fig 6A). Almost all errantiviruses that sit within the “insect errantivirus” group have duplicated the RNase H domain similarly to vertebrate retroviruses, with the C terminal copy presenting itself as a “pseudo” RNase H domain (Fig 6A-B). This “pseudo” RNase H domain retains the barrel-like structure made of five core β-sheets, but the order of the second and the third β-sheets is flipped (Fig 6A).

On the other hand, “ancient errantiviruses” carry an RNase H domain, and a C-terminal “mini” domain. This consists of three small core β-sheets and α-helix residues at its N terminus. This “mini” domain is also carried by insect elements in this group, further suggesting that those insect elements are equally ancient as the other errantiviruses in this group. Furthermore, this “mini” domain is also found in the bridge region of yeast S. pombe tf1 pol (Fig 6B-C), indicating it as a feature of highly ancient transposons. Thus, Figure 6 links the POL extended RT/connection phylogenetic separation of insect and ancient errantiviruses with a structural difference in the bridge region of Pol.

The other structures (“Mini” domains 2, 3 and “Pseudo” domain 2) present themselves less frequently as outliers, and are restricted to specific clades, or even solitary elements (highlighted in yellow and dark blue in Fig 6B). One notable outlier is an insect element within the “ancient errantivirus” group that carries a variation of a “pseudo” RNase H domain (position 7, noted as “pseudo” domain 2 in Fig 6). This same domain can be found in a phylogenetic group in the “insect errantivirus” group (position 3 in Fig 6B). Furthermore, some elements in this “insect errantivirus” group carry a “mini” domain as well (position 3 in Fig 6B). Another is that some elements within the “insect errantivirus” group only carry a solitary RNase H domain (positions 4 and 5 in Fig 6B). It should be noted that these outliers are mostly specific to small phylogenetic groups/clades, and likely also arose anciently.

Intact errantiviruses are sourced from various lineages of Ty3/gypsy retrotransposons

To further investigate the origins of env-containing errantiviruses, we assembled a phylogenetic tree from the extended RT/connection region of representative errantiviruses and included non-env-containing retrotransposons that fall under the broader umbrella of “Ty3/gypsy” elements. This includes transposons such as MAG and Tor, and Chromoviruses such as Maggy, which are annotated in the GypsyDB (https://gydb.org). We expected that env-containing errantiviruses would form a monophyletic clade in the tree if they arose from a specific non-env-containing Ty3/gypsy element.

We firstly find that most non-env-containing Ty3/gypsy elements sit in the ‘ancient errantivirus’ clade, with the exception of the 412/MDG1 group – specifically the Drosophila gypsy elements BLOOD and Gypsy-28_Dan – which cluster with Drosophila and other insect errantiviruses, consistent with a previous study (Bargues and Lerat 2017) (Fig 7). Additionally, we find that the non-env-containing elements rarely regularly form discrete clades without env-containing errantiviruses. For example, Tor4b is found sister to Tunicata elements carrying F-type env, and Cer1 was found sister to a Platyhelminthes errantivirus carrying HSV/gB-type env. Interestingly, Chromoviridae, which are found in Fungi and Plants and are representative of highly ancient, widespread Ty3/gypsy elements (Gorinsek et al. 2004; Kordis 2005), are found to cluster within the tree, nested with “ancient” errantivirus clades (Fig 7). They sit within a clade that includes Tunicata errantivirus carrying HSV/gB-type env.

Trees of known Ty3/gypsy elements and newly identified Errantiviruses.

Midpoint-rooted maximum-likelihood tree of the POL extended RT/connection region from representative env-containing errantiviruses and selected non-env-containing Ty3/gypsy retrotransposons. Representative non-env-containing Ty3/gypsy elements from animals, fungi and plants are included to place the newly identified errantiviruses within the broader diversity of Ty3/gypsy retrotransposons. Host Order and env type given for each clade, as well as solitary elements on the tree are shown. Errantiviruses containing chromodomains are marked in yellow. The “ancient errantivirus” group is coloured in green and “insect errantiviruses” in black. Known Ty3/gypsy elements includes Gypsy Group 1 OSIRIS (Gypsy-21_DAn), Gypsy Group 1 OSVALDO (Dypsy-3_DEl), Gypsy Group 2 BICA (Gypsy-6_DFi), Gypsy Group 2 BLASTOPIA (Gypsy-32_DWil), Gypsy Group 2 MDG3 (INVADER3, MDG3_DM), Gypsy Group 2 MICROPIA (Gypsy-7_DSe), Gypsy Group 2 SACCO (Gypsy-10_DAn), Gypsy Group 3 412/MDG1 (BLOOD, Gypsy-28_Dan), Gypsy Group 3 CHIMPO (Gypsy-7_DAn), Gypsy Group 3 17.6 (DM176, IDEFIX, ZAM), Gypsy Group 3 GYPSY (Gypsy-5_Dgri, Gypsy-Gypsy), Mag (Mag_As, MAG_Bmor), Cer (Cer1, Cer4), Tor (Tor1, Tor2, Tor4b), Athila/Tat (ATHILA0p1), and Chromoviridae (ATHILA0p1, CRM_AAM94350.1, REINA, Saccharomyces_Ty3, SCL213_Galadriel, Skipper*, Spombe_Tf1_AAA35339.1, SUSHI).

The annotated Ty3/gypsy elements present themselves in the tree, spread between and within different “ancient” errantiviruses from different hosts, with little pattern to their spread regarding host or env type. Notably, a clade of errantiviruses from Cnidaria, Mollusca and Annelida hosts all carry a chromodomain, a chromatin-binding domain found C-terminal to Integrase in chromoviruses and some related Ty3/gypsy elements. This clade sits basally to known chromoviruses, but sister to non-chromodomain-containing errantiviruses (Fig 7, Fig. S9). We also observe outliers, where only one or two errantiviruses carry chromodomains within the whole clade. For example, Annelida_Polychaeta_Dgyr_Errantivirus_1 contains a chromodomain, but is located in a clade with non-chromodomain-containing errantiviruses (Fig. S9). These observations rule out a single origin of env-containing errantiviruses and instead suggest that multiple lineages of ancient Ty3/gypsy elements contributed to the current distribution and diversity of errantiviruses.

Discussion

We found copies of env-containing Ty3/gypsy retrotransposons (errantiviruses) in the genomes of nearly every major metazoan order. They are intact, carrying uninterrupted ORFs of all three retroviral proteins and are flanked by LTRs. Additionally, most errantiviruses identified in this study have multiple copies in their respective genomes, consistent with recent or ongoing genomic expansion. When we further characterized these elements, we found that they carry different ENV proteins; ENV resembling glycoprotein F from paramyxoviruses, and ENV resembling glycoprotein B from herpesviruses. Despite great divergences in the primary sequence, the structural domains of these proteins are highly conserved. Throughout this manuscript, we have used “errantivirus” as an operational term for env-containing Ty3/gypsy retrotransposons, including elements outside Drosophila. This usage should not be interpreted as a formal taxonomic proposal, nor as evidence that all such elements share the same amplification mechanism as Drosophila gypsy. Rather, it reflects their shared genomic organisation, Ty3/gypsy ancestry and possession of env-like fusogen genes.

We found that these errantiviruses are highly ancient, being found in animals such as cnidarians, ctenophores and tunicates – lineages that diverged early in metazoan evolution. We identified an “ancient errantivirus” group, which sits separately to elements found in insects (Fig 2). These ancient elements are also structurally distinct, with the bridge region between RT and IN of pol resembling that of yeast Ty3 retrotransposons, rather than that of insects. Additionally, pol genes of non-env-containing Ty3/gypsy elements from diverse eukaryotic taxa form clades with the “ancient errantivirus” group (Fig 7). Although these patterns are consistent with their ancient nature, the specific origins of env-containing errantiviruses in metazoans remain elusive. Errantiviruses from the same or closely related phyla tend to cluster together in the POL extended RT/connection phylogenetic tree (Fig. S4). However, the pattern becomes convoluted when highly ancient errantiviruses are compared, as elements from distantly related phyla often cluster together without representation from intermediate animal lineages (Fig. S4 and Fig 7). Many of these clades tend to have longer branches, suggesting that they are distantly related and diverged a long time ago. For example, within tree position A10 (Fig. S4), a clade consisting of elements from Hexanauplia and Annelida are separated by longer branch lengths. The reason for these types of clades, and as to why we do not see elements in species between those distantly related orders, is likely due to the frequent loss of these retrotransposons from the genome. In such cases, distantly related species may retain intact or recently expanded copies that were vertically inherited and diverged with the host species, while copies in other species may have become inactive and degraded over time.

The errantiviruses we identified form multiple clades with previously characterized non-env-carrying Ty3/gypsy elements in the pol tree (Fig 7). Additionally, some errantiviruses carry chromodomains, but do not always cluster together with other chromodomain carrying retrotransposons. This pattern could be explained by different lineages of Ty3/gypsy frequently acquiring env from other errantiviruses. This would explain annotated Ty3/gypsy elements forming clades with both HSV/gB– and F-type env-containing errantiviruses. Another explanation could be that widespread horizontal transfer is occurring between distantly related species. As the species diverge from each other, different types of env could be acquired, lost and regained multiple times. However, neither of these explanations account for the congruence seen between pol and the host, at least at the level of animal classes and orders (Fig. S4), nor the structural divergence of HSV/gB-type ENV (Fig 5 and S8). The spread of errantiviruses is indicative of multiple lineages of retrotransposon contributing to the current picture of errantiviruses, but the congruence supports their ancient nature. One plausible explanation is that different lineages of Ty3/gypsy within the last common ancestor acquired env, which were inherited and diverged with the host species, infrequently recombining with other elements within the same/similar hosts. The non-canonical GAG-ENV-POL arrangement further supports this view. These elements are found in several non-insect animal groups and do not form a single monophyletic clade, indicating that this arrangement is not simply a recent derivative of one lineage (Fig. S4 and S8). Instead, ORF rearrangement appears to be another feature that has diversified among long-standing HSV/gB-type errantivirus lineages.

Our interpretation does not exclude lineage-specific loss or rare horizontal transfer. Indeed, extensive loss is expected for ancient retrotransposon lineages, because inactive copies will progressively degrade and disappear from detectable sequence space. Similarly, occasional horizontal transfer between closely related hosts may contribute to local discordance. To examine whether host-taxonomic structure is nevertheless retained at local evolutionary scales, we performed targeted host-label permutation analyses of selected Annelida and Lepidoptera POL RT subclades (Fig. S5). The Annelida subclade showed strong element–host concordance at both host-family and host-species levels. The Lepidoptera subclade showed a more mixed pattern, with several highly supported clades enriched for related host groups at the superfamily or broader taxonomic level, but also a substantial number of mixed clades. Thus, lineage-specific loss and rare horizontal transfer are compatible with our data, but frequent broad horizontal transfer between distantly related animal groups is unlikely to explain the overall host-taxonomic structure.

Acquisition of ENV proteins by errantiviruses

It was originally speculated that errantiviruses acquired env from baculoviruses in insects, some of which carry an F-type ENV protein Ld130, potentially via recombination events (Malik et al. 2000; Rohrmann and Karplus 2001). While Ty3/gypsy elements may have obtained the env gene via recombination, it is unlikely that baculoviruses are the source for the env gene found in errantiviruses, as they are far more widespread than insects. The congruence of phylogeny between env and pol indicates that they have likely co-evolved (Fig 4). This, along with the presence of errantiviruses in ancient host orders such as Cnidaria, Ctenophora and Tunicata, suggests that env-carrying Ty3/gypsy elements are highly ancient transposons, rather than the result of a recent recombination event with an exogenous virus. HSV/gB-type env in errantiviruses appears to be equally ancient, as it is similarly found in ancient animal orders (Fig 1). The arrangement of cysteine bridges that stabilize the structure of the ectodomain, are consistent between elements within the same order, even though pol presents itself separately on the tree (Fig 5 and S8). This implies that the env has diverged with the host but likely undergoes recombination events with other elements within the same species. Though, pol genes with HSV/gB-type env mostly cluster within their own clades and are not mixed with pol with F-type env, indicating that recombination events are not happening frequently enough to randomise their spread, and likely occurred anciently. The origins and acquisition, therefore, of either type of env, likely occurred in the last common ancestor of metazoans.

The survival mechanism of errantiviruses in the genome

In Drosophila, errantiviruses occupy the somatic cell niche in the ovaries that surround the germline, and the expression of env allows virus-like particles to infect the germline to create new copies (Song et al. 1994). As an adaptation, Drosophila species have duplicated the germline transposon defence mechanism to target errantiviruses in the somatic cells of the gonad (Brennecke et al. 2007). We recently showed that errantiviruses are targeted by piRNAs in ovarian somatic cells across multiple insect orders, including mosquitoes, bees and crickets (Belot et al. 2020). It is therefore possible that this is more universal, and that env-containing retrotransposons occupy the cell niches surrounding the germline, and that expression of env allows them to continue to infect the germline. For example, murine IAP retrotransposons that carry env genes also remain infectious, like Drosophila errantiviruses (Ribet et al. 2008). However, these additional ORFs and genes in endogenous retroviral elements are constantly subject to evolutionary pressure from their hosts (Wang and Han 2020; Wang et al. 2020). It is therefore significant that the env gene is as highly conserved as we have observed, and that it does not seem to have been subject to frequent loss and acquisition. It is also important to note the mechanisms of transmission via the action of this env gene remain elusive for most of the elements we identified, except for those in Drosophila species. However, the structures formed by env, for both HSV/gB and glycoprotein F, form the same functional structures as those found in exogenous viruses, required for cell entry (Heldwein et al. 2006; Connolly et al. 2011; McLellan et al. 2013). While entry of HSV also requires three other glycoproteins, gD, gH and gL, gB is specifically required for membrane fusion (Johnson et al. 2011; Jambunathan et al. 2021). The mechanism of cell entry, therefore, is unknown for these errantiviruses that carry gB, and necessitates further investigation regarding their expression. Though, given their highly structured configuration, it is likely that they retain membrane fusion capabilities. Additionally, given the widespread nature of these env-containing elements, it may also be possible that somatic transposon defence mechanisms against them are more common.

In summation, the env gene appears to have been anciently acquired by Ty3/gypsy elements, and has been retained by those elements, diverging as their host species evolved. Today, we see a snapshot of errantiviruses widespread throughout metazoan genomes, continuing to create new insertions to survive long term.

Materials and methods

Genome assemblies used in the study

Reference genome sequences were retrieved from NCBI and grouped into the following insect orders and other metazoan phyla: Diptera, Lepidoptera, Coleoptera, Hymenoptera, Hemiptera, Polyneoptera, and Palaeoptera for insects; Arachnida, Branchiopoda, Chilopoda, Collembola, Diplopoda, Hexanauplia, Malacostraca, Merostomata, Ostracoda, and Thecostraca for Arthropoda; Acanthocephala, Annelida, Brachiopoda, Bryozoa, Dicyemida, Mollusca, Nemertea, Onychophora, Orthonectida, Phoronida, Platyhelminthes, Priapulida, Rotifera, and Tardigrada for Protostomia; Cephalochordata, Echinodermata, Hemichordata, and Tunicata for Deuterostomia; Cnidaria, Ctenophora, and Porifera. We obtained lists of reference genomes in each category in May 2023 using the Entrez Direct command line tool. The list of the genome assemblies used can be found in Supplementary Table S6.

Identification of intact genomic copies of Ty3/gypsy errantiviruses

We ran a total of five iterations of tBLASTn (ncbi-blast-2.9.0) to search for the genomic copies of Ty3/gypsy errantiviruses, firstly using the baits of known elements, and subsequently using the elements that were identified after each iteration. We chose new baits that were not represented in the phylogenetic tree of the POL (see below for the detail) from the previous iteration and repeated the process until we no longer found new elements. We used GAG, POL and ENV protein sequences of the Dipteran gypsy errantiviruses from the RepBase (https://www.girinst.org/repbase/) for the first tBLAST search. In this search pipeline, “ENV” is used as a generic term for candidate envelope-like proteins. Candidate ENV ORFs were classified after identification as either F-type ENV or HSV/gB-type ENV based on sequence similarity, domain prediction and structural features. The initial ENV baits were derived from Dipteran gypsy errantiviruses, whereas subsequent iterations also used ENV proteins from newly identified representative elements, allowing both F-type and HSV/gB-type ENV-containing elements to be recovered. The bait sequences for all tBLASTn iterations can be found at https://github.com/RippeiHayashi/errantiviruses_2025/tree/main. We ran tBLASTn with default parameters (-evalue 10). We applied different criteria to identify ERVs from species outside insects as we did not have prior knowledge of whether or what kind of ERVs are found in those species. We briefly summarise our approach below.

Genomic regions that showed homology to the POL baits with a bit score of greater than 50 or to the GAG and ENV baits with a bit score of greater than 30 were individually pooled. Genomic regions that had hits from multiple baits were merged for their coordinates using bedtools/2.28.0. Regions that were larger than 1500 nucleotides for POL and 300 nucleotides for GAG or ENV were collected. Following, for insect genomes we looked for regions that carry GAG, POL, ENV sequences that are next to each other in this order allowing up to 500 nucleotides of gaps between them as well as excluding regions of smaller than 4000 nucleotides or larger than 10000 nucleotides. For the other genomes we allowed up to 1000 nucleotides of gaps between POL and ENV and collected regions that had both ORFs that are not smaller than 3000 nucleotides or larger than 10000 nucleotides.

We subsequently checked whether the regions contain uninterrupted ORFs by making translated sequences using transeq from EMBOSS-6.6.0 and excluding ORFs with more than five stop codons. We clustered all the identified GAG, POL and ENV from each genome by Clustal Omega (https://www.ebi.ac.uk/Tools/msa/clustalo/) and identified genomic regions that harbour all three ORFs that are multiplied in the genome. We also searched for ORFs regardless of the copy number when there were no multiplied elements in the genome. When genomic copies were abundantly found in multiple closely related species, we only annotated the elements from representative genomes. Genomes that were annotated and the number of predicted ORFs per genome are summarised in Supplementary Table S6.

These length, gap and stop-codon thresholds were used as permissive prefilters to recover candidate elements for downstream curation, rather than as the sole definition of intact errantiviruses. We chose inclusive thresholds at this stage to avoid excluding divergent elements, elements with non-canonical ORF arrangements, or elements with small annotation errors. Final inclusion required additional evidence, including Ty3/gypsy POL classification, recovery of full-length RT and Integrase domains within continuous ORFs, HHpred-supported domain annotation of GAG, POL and ENV, and removal of elements with large domain truncations.

The above process yielded abundant putative intact copies of ERVs from insect genomes whereas we took additional steps for the other genomes. For Arthropoda genomes outside insects as well as other Protostomia genomes, we extended the genomic regions of predicted GAG-POL to downstream by 3000 nucleotides and manually annotated ENV ORFs using Benchling (https://www.benchling.com/). The identity of GAG ORFs were further confirmed by HHpred (https://toolkit.tuebingen.mpg.de/tools/hhpred). For Cnidaria, Ctenophora and Deuterostomia genomes outside vertebrates we used transeq to annotate putative GAG and ENV ORFs from 5000 nucleotides downstream and 3000 nucleotides upstream to the predicted POL ORFs, respectively.

We took confident 945 POL ORFs from the first tBLASTn search and made a multiple sequence alignment of the conserved regions of the Integrase domain using mafft-7.505 with the option ‘--auto’, and inferred the phylogenetic relationships using iqtree-1.6.12 with the option ‘-m rtREV+R4 –bb 1000’ (see more details in the section below). We took 152 representative POL from the phylogenetic tree and used the POL and the corresponding GAG and ENV ORFs from the same genomic regions for a second tBLASTn search. We repeated the above steps after combining the results of the two tBLASTn searches. We did not find any ERVs in Nematoda genomes that have GAG, POL and ENV in this order while the previous study reported non-canonical Nematoda ERVs that have flipped the order of ENV and POL (Sacco et al. 2022). Thus, we searched for regions that have ENV and POL in this order allowing gaps of up to 2000 nucleotides between them. GAG was subsequently predicted from 5000 nucleotides upstream to the predicted ENV-POL. We extended the search for potential GAG-ENV-POL insertions to other non-insect genomes and found similar elements in Bryozoa and Platyhelminthes. We also found ENV-GAG-POL elements in Arachnida genomes (Medioppia subpectinata and Oppia nitens). These elements were recovered through the search of ENV-POL elements because GAG ORF was short. We did not extend our search to find other arrangements of ORFs.

Additionally, we found that the glycoprotein gene present in mosquito BEL/pao retrotransposons (Dezordi et al. 2020) are also found next to gypsy POL genes. The glycoprotein gene resembles the envelope genes from Chuviridae and is structurally similar to glycoprotein B from Herpesviruses. We identified ten mosquito BEL/pao retrotransposons that carry this glycoprotein from RepBase (BEL-199_AA, BEL-279_AA, BEL-652_AA, BEL-801_AA, BEL-804_AA, BEL-879_AA, BEL-51_AnFu, BEL-56_AnFu, BEL-81_AnFu, and BEL-71_CQ) and used them as baits for a tBLASTn search against Arthropoda genomes. The tBLASTn search identified 114 gypsy elements with homologous glycoprotein genes from Coleoptera, Hymenoptera, Polyneoptera, and Hexanauplia species. Combined together, the second iteration of tBLASTn search identified 451 new elements. The phylogenetic analysis of POL from the elements identified in the first and second tBLASTn runs revealed clades that only contained new elements from the second tBLASTn run, suggesting that the first two tBLASTn runs were not comprehensive. Therefore, we took 34 additional baits to cover all the major clades in the phylogenetic tree for a third tBLASTn search, which identified 61 new elements. We repeated tBLASTn runs five times in total; the fourth and fifth runs used 17 and 7 new baits and identified 49 and 6 new elements, respectively. We stopped the tBLASTn search after five iterations because the phylogenetic analysis of POL did not reveal new clades.

Following, we further curated newly identified ERVs. Firstly, we used hmmer 3.3.1 (http://hmmer.org/) to make sure that the ERVs that we identified are classified as Ty3/gypsy. We compared POL ORFs from the newly identified ERVs against the POL ORFs of known retroelements that are available from the Gypsy Database 2.0 (https://gydb.org/index.php/Main_Page). The Gypsy Database 2.0 collects hmm profiles of Ty1/Copia, Bel/Pao, Ty3/gypsy, and Retrovirus POL proteins. The hmmer search classified all the newly identified ERVs into Ty3/gypsy. Secondly, we ran HHpred from hh-suite 3.3.0 (Steinegger et al. 2019) or the online tool of HHpred (https://toolkit.tuebingen.mpg.de/tools/hhpred) to predict GAG, ENV, and POL ORFs using the default search parameters. For the command line HHpred, we first made multiple sequence alignments using hhblits and the Uniclust database UniRef30/2022_02 (https://uniclust.mmseqs.com/). This was followed by hhsearch to predict domain structures using the collection of pdb70 from March 2022 (https://www.user.gwdg.de/∼compbiol/data/hhsuite/databases/hhsuite_dbs/). We used default settings to run hhblits and hhsearch. ENV and POL ORFs with large truncations were removed after the domain structure prediction. ERVs that predicted all GAG, ENV and POL ORFs were taken for further analyses.

We additionally ran BLASTn using the genomic regions that cover all three ORFs as baits to interrogate the copy number of ERVs in the genome where they are found. The results are summarised in Supplementary Table S1 where we counted the number of BLASTn hits with a more than 98% identity in more than 98% of their length to the bait sequences. A majority of newly identified ERVs (1386 out of 1512 elements) carry ORFs that are found at least twice in the genome. These copy-number estimates were used as evidence consistent with recent or ongoing genomic expansion, but were not taken as direct evidence of current transposition activity.

For selected Porifera and Echinodermata genomes in which no intact GAG-POL-ENV elements were detected, we additionally searched for full-length GAG-POL Ty3/gypsy elements to assess whether intact retrotransposons could be recovered from these assemblies. Two representative genomes were analysed from each phylum. Candidate regions containing GAG and POL in the expected order were identified and curated using the same general criteria applied to GAG-POL-ENV elements, except that no downstream ENV ORF was required. Candidate GAG-POL elements were then assessed for copy number, putative LTRs and tRNA primer-binding sequences as described above.

Identification of putative long terminal repeats and transfer RNA (tRNA) primer binding sequences

Long terminal repeats (LTRs) of ERVs were identified as follows. Firstly, we fetched the genomic coordinates of ERV insertions and extended 3000 nucleotides at either ends from the Start codon of GAG and the Stop codon of ENV ORFs (POL was taken for the GAG-ENV-POL ERVs) for all species categories except for Cnidaria, Ctenophora and Deuterostomia where we extended 5000 nucleotides at either ends from the Start and Stop codons of POL. We retrieved the genomic sequence of the extended ERVs and ran BLASTn against the same sequence to identify repeated sequences. We considered repeated regions of greater than 99% identity and the tBLASTn bit score between 200 and 5000 as putative LTRs. We excluded LTRs whose distance between them is shorter than 5000 nucleotides or longer than 14000 nucleotides. Following, we searched for putative tRNA primer binding sequences (PBS). tRNA sequences were retrieved from genomic annotations from representative genomes in respective species categories except for Ctenophora where we took tRNA sequences from Protostomia, Cnidaria and Deuterostomia because there were no annotated genomes available at the time of the study. Representative genomes that we used to retrieve tRNA sequences are: GCF_000001215.4 and GCF_013141755.1 for Diptera, GCF_014905235.1 and GCF_905147365.1 for Lepidoptera, GCF_000002335.3 and GCF_917563875.1 for Coleoptera, GCF_003254395.2 and GCF_016802725.1 for Hymenoptera, GCF_014356525.2 and GCF_020184175.1 for Hemiptera, GCF_002891405.2 and GCF_021461395.2 for Polyneoptera, GCF_921293095.1 for Palaeoptera, GCF_000671375.1, GCF_020631705.1, GCF_002217175.1, GCF_016086655.3, GCF_024679095.1, GCF_019059575.1 for Arthropoda, GCF_000326865.2, GCF_020536995.1, GCF_000002985.6, GCF_000715545.1 for Protostomia, GCF_000222465.1 and GCF_022113875.1 for Cnidaria, GCF_013122585.1, GCF_000003605.2, GCF_000003815.2, and GCF_000002235.5 for Deuterostomia. The minimum requirements of the priming events are incompletely understood, but the length of PBS for retrotransposons is generally between 8 and 18 nucleotides and the PBS is positioned near the 3’ end of the 5’ LTR (Wilhelm and Wilhelm 2001). We chose the 12 nucleotides at the 3’ end of tRNAs as putative primer sequences and searched for PBS in the 100 nucleotide-window centred at the 3’ end of the 5’ LTR using bowtie-1.2.3 allowing up to two mismatches. We added “CCA” to the 3’ end of the tRNA sequence when it is not present in the genomic sequences. We successfully identified putative PBSs in 1151 out of 1512 ERVs that we identified in this study. We grouped ERVs based on the type of tRNA used for the PBS and summarised the results in Figures S2 and S4 as well as Supplementary Table S1.

Phylogenetic analysis of POL

We obtained multiple sequence alignments of the POL extended RT/connection region and Integrase domains as follows. For clarity, we distinguish between the extended RT/connection region used for phylogenetic analysis and the RNase H-containing bridge region analysed structurally in Figure 6. The extended RT/connection region comprises the RT polymerase core, formed by the fingers, palm and thumb subdomains, together with the downstream connection subdomain. Although the connection subdomain is included as part of retroviral RT in structural studies, it corresponds to a partial RNase H-like fold and has been interpreted in evolutionary studies as a degenerated RNase H-like tether domain (Malik and Eickbush 2001). In contrast, the bridge region analysed in Figure 6 is defined here as the region spanning the canonical RNase H domain and the C-terminal region between RNase H and Integrase. In the structural scheme used in Figure 6A, an intact RNase H domain consists of β1–β2–β3–α–β4–α–β5–α, whereas the connection/tether region included in the extended RT alignment corresponds only to the partial RNase H-like fold β1–β2–β3–α–β4–α.

For the POL extended RT/connection region, we first took the HHpred alignments of POL against the RT structures of Moloney Murine Leukaemia Virus (Protein Data bank entry 4MH8_A) and Human Immunodeficiency Virus type 1 (Protein Data bank entry 4G1Q_B), spanning regions that correspond to their fingers, palm, thumb and connection subdomains (Das and Georgiadis 2004), namely the amino acid positions of 28 to 426 and 7 to 405, respectively. Thus, the region used for this phylogenetic analysis is broader than the RT polymerase core alone, but does not correspond to the complete canonical RNase H domain analysed structurally in Figure 6. This region was used because it is conserved and alignable across the errantivirus POL proteins analysed here. However, this process caused small truncations at either ends of the domain because of poor alignments at the individual substructures. To overcome this issue and to facilitate sequence alignments, we extended the N– and C-termini by 10 and 15 aa, respectively, after merging the two predicted regions of RT domains. We then ran mafft-7.505 to generate a multiple sequence alignment, from which we manually trimmed the either ends to have intact fingers, palm, thumb and connection subdomains for all POL ORFs. Similarly, we took HHpred alignments of POL against the Integrase domains of Human Spumaretrovirus (Protein Data bank entry 3OYM_A) and Simian T-lymphotropic Virus type 1 (Protein Data bank entry 7OUH_B), spanning regions that correspond to their N-terminal domain (NTD) and catalytic core domain (CCD) (Engelman and Cherepanov 2014), namely the amino acid positions of 54 to 273 and 8 to 197, respectively. We used the predicted Integrase domains with 15 and 30 aa extensions at the N– and C-termini, respectively, for generating a multiple sequence alignment before manually trimming the ends. We inspected the multiple sequence alignments using Mview (https://www.ebi.ac.uk/Tools/msa/mview/) and removed ERVs that contained truncated RT or Integrase domains from the analyses. We included ORFs from 5 Drosophila errantiviruses (DM176, IDEFIX, ZAM, GYPSY, and Gypsy-5_DGri) in the multiple sequence alignments.

We ran iqtree-1.6.12 with the option ‘-m rtREV+R4 –bb 1000’ to generate the phylogenetic trees of all POL Integrase domains and extended RT/connection regions, and those from ERVs found in vertebrate and invertebrate genomes, respectively. We then selected representative elements from clades, taking into consideration the relative distances between nodes in the same clade. For example, if the distance between two nodes is less than 0.1, meaning that there is little – no variations between their sequences, only one node in that position is taken. Additionally, we made sure that the representatives captured the diversity of env found within the clade. For example, if a clade contains both F-type and HSV/gB-type elements, at least one for both is taken, regardless of distance. Furthermore, the phylogeny of the hosts was represented as fully as possible. For example, if a clade has elements from Diptera and Coleoptera, at least one from both is taken regardless of distance. Once representatives were collected, the tree was then accordingly pruned. It should be noted that the final pruned tree of representatives was only used as a figure aid for clarity (Fig 1-4), and all analyses were done on the full Integrase and extended RT/connection phylogenetic trees. We used a similar approach as detailed to select representatives for structural analysis of env ectodomains, as well as the pol bridge region (Table S4). If any outliers were found, remaining elements from the same clade were also analysed, to check whether structural inconsistencies were consistent across the clade.

For congruence plots in figures 4 and S2, we grouped elements based on which node they sit in the tree, ensuring that the node is well supported by bootstrap values greater than 95. We used the networkD3 package in R to generate Sankey plots to visualise the representation of elements from one tree in another. The node definitions are summarized in Table S5.

Further, we took 86 representatives of the F-type ENV based on their positions in the corresponding POL phylogenetic tree for the analysis. We identified the signal peptides and transmembrane regions by DeepTMHMM (https://dtu.biolib.com/DeepTMHMM) and obtained the ectodomains that are in between them. We subsequently ran mafft-7.505 to generate a multiple sequence alignment. To compare errantivirus F-type ENV proteins with homologous fusogens outside Ty3/gypsy retrotransposons, we generated an additional F-type ENV ectodomain alignment that included representative external sequences. These included baculovirus F proteins, paramyxovirus and pneumovirus F proteins, and the BEL/Pao-associated Drosophila melanogaster Roo F-like protein.

Host-taxonomic concordance analyses

For targeted host-taxonomic concordance analyses, we selected representative Lepidoptera and Annelida POL extended RT/connection subclades from tree positions I1 and A6, respectively, and scored internal nodes with bootstrap support ≥95 and 3–10 descendant tips. Host-taxonomic labels were randomly permuted across the fixed POL extended RT/connection topology 10,000 times while preserving the observed label counts, and the same supported nodes were rescored after each permutation. In the Lepidoptera analysis, family labels were permuted and superfamily assignments were derived from the permuted labels. In the Annelida analysis, host-species labels were permuted and host-family assignments were derived from the permuted labels.

Structure prediction of ENV proteins, the RNase H and chromodomains using Alphafold 2

We took representatives of the F-type and HSV/gB-type ENV proteins and the RNase H-containing bridge region, defined here as the canonical RNase H domain and the C-terminal region between RNase H and Integrase, using branch definitions from the phylogenetic trees of POL extended RT/connection region. We also took the C-terminal half of POL, including the Integrase domain, from representative errantiviruses, investigated in Figure 7. We predicted their structures when chromodomains were detected by hhpred (see above). We used the colabfold wrap v1.4.0 of Alphafold 2 (https://github.com/YoshitakaMo/localcolabfold/releases/tag/v1.4.0), and used MMSeqs2 for the database search and clustering of protein sequences before performing the structural prediction (Mirdita et al. 2022). We used the whole ectodomains (see above) of the F-type and HSV/gB-type ENV proteins and predicted the trimer structures using the multimer mode of Alphafold 2 (AlphaFold2-multimer-v2) or Alphafold server (https://alphafoldserver.com/). We then visualised and characterised visible structures and disulfide bonds or cysteine bridges using Maestro (Schrödinger Release 2024-3) and grouped them by common substructures (Table S4). Structures of all predicted chromodomains in POL were confirmed by Alphafold. Ectodomains of viral class III fusogens of herpesvirus gB, rhabdovirus G, orthomyxovirus GP75/GP64-like proteins, baculovirus GP64, BEL/Pao-associated gB-like proteins were also included in this analysis when experimentally resolved structures were not available. Further, pairwise amino-acid identities were calculated within each viral family or errantivirus HSV/gB-type ENV group using curated ectodomain alignments. For each pairwise comparison, alignment columns containing a gap in either sequence were excluded, and identity was calculated as the number of identical residue–residue positions divided by the total number of compared residue–residue positions. These gap-excluded pairwise identities were used as descriptive measures of primary-sequence divergence and were not interpreted as evolutionary distances or molecular-clock estimates.